### 2023-03-27 paper/paper.tex
init paper

\pdfoutput=1

\documentclass[11pt]{article}

\usepackage[]{acl}
\usepackage{tabularx}

\usepackage{times}
\usepackage{latexsym}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage{microtype}

\title{Cross-linguistic complexity analysis using UD treebanks}

\author{Aron Winkler \\
  University of Tübungen \\
  \texttt{email@domain}}

\begin{document}
\maketitle
\begin{abstract}
    The aim of this project is to explore the use of UD treebanks in cross-linguistic complexity research. While it is not this paper's objective to make strong theoretical claims on complexity in its own right, it examines complexity measures computed on various UD treebanks across 4 languages. This work focuses on treebanks for English, German, Hungarian, and Chinese, in an attempt to determine whether complexity measures stay consistent within treebanks for the same language, and whether some languages consistently show higher complexity than others.
    These claims are verified with official UD corpora, as well as with a parallel corpus of 50 sentences for each language mentioned above.
\end{abstract}

\section{Introduction}

What constitutes linguistic complexity is an elusive multi-faceted inquiry, and quantifying it is equally challenging. Previous approaches have included text-based metrics as well as eye-tracking studies (e.g. \citealp{Lee:2007}), but a general consensus has not been reached, especially when it comes to the notion of "overall complexity" of a language. In the case that an experiment produces numeric values as a proxy for complexity, it is not obvious that results computed on two different languages are comparable.

In this work, the objective is not to attempt to resolve these foundational question about linguistic complexity, but rather to accept that there are certain well documented text-based proxies for computing it - such as lexical, morphological, and syntactic complexity measures - and to explore how the Universal Dependencies framework can be employed to further the subject matter.

Indeed, the potential of UD-annotated data appears to be largely untapped in linguistic complexity research. A convincing contribution to this field was made by \citealp{berdicevskis-etal-2018-using}, estimating robustness of complexity values computed on a variety of treebanks, while accounting for additional obstacles such as language-specific annotation conventions that might reduce comparability.

While this project is more restricted in scope and claims, the goal is to showcase the use of UD treebanks in linguistic complexity research along the same lines as \citealp{berdicevskis-etal-2018-using}. For this purpose, a variety of treebanks across four languages were used, in combination with a set of text-based complexity metrics inspired in part by readability research, and in part by complexity research more broadly. These metrics are computed on two datasets, one of which is a collection of official UD treebanks, and the other being a small in-house parallel corpus comprising 50 sentences of each of the 4 target languages.

\section{Dataset and metrics}

\subsection{Dataset}

\input{tables/webdata.tex}

Table~\ref{tab:webdata} shows the list of UD web treebanks used in this experiment. For the English language, which has many treebanks available to it, the Atis treebank (Github\footnote{\url{https://github.com/UniversalDependencies/UD_English-Atis}}), the gum GUM (\citealp{berzak2016tle}, GitHub\footnote{\url{https://github.com/UniversalDependencies/UD_English-GUM}}), and the EWT treebank (\citealp{silveira14gold}, GitHub\footnote{\url{https://github.com/UniversalDependencies/UD_English-EWT}}) were ised. For German, this work employs GSD (\citealp{mcdonald-etal-2013-universal}, GitHub \footnote{\url{https://github.com/UniversalDependencies/UD_German-GSD}}) and HDT (\citealp{borges-volker-etal-2019-hdt}, \citealp{hennig-kohn-2017-dependency}, \citealp{hdtextra}, \citealp{hdtguide}, GitHub \footnote{\url{https://github.com/UniversalDependencies/UD_German-HDT}}). For Chinese, 
GSD (GitHub\footnote{\url{https://github.com/UniversalDependencies/UD_Chinese-GSD}}) and PUD (GitHub\footnote{\url{https://github.com/UniversalDependencies/UD_Chinese-PUD}}) were the final choice. For Hungarian, only the Szeged treebank (\citealp{szeged}, Github\footnote{\url{https://github.com/UniversalDependencies/UD_Hungarian-Szeged}}) was available - therefore in order to have at least two treebanks for each language, the sentences annotated in the Szeged treebank were randomly distributed across two new treebanks (hun\_szeged\_1, hun\_szeged\_2).

An additional parallel treebank\footnote{\url{https://github.com/iscl-dtdp/ParallelTreebank-FinalProject}} made by Nino Meisinger, Qin Gu, Lisa Wang, and Aron Winkler as coursework at the University of Tübingen was also employed. Although its limited size of only 50 sentences per language prevents meaningful conclusions from being drawn from measurements computed on it, its parallel nature nevertheless allows interesting observations to be made in reference to the complexity metric values obtained from the official treebanks. Sentences for this parallel treebank were sourced from Tatoeba\footnote{\url{https://tatoeba.org/en/}}, more specifically by selecting the 50 longest sentences that had translations for all 4 target languages. If more than one translation was available for any given language, then the choice was left up to the annotator. The final treebank was then manually annotated according to UD standards by annotators native or very proficient in the target language.

\subsection{Metrics}

\input{tables/metrics.tex}

Table~\ref{tab:metrics} details the complexity metrics used in this experiment. Most of them were inspired by readability research, while a couple were adopted from \citep{berdicevskis-etal-2018-using}. Sentence-level metrics (labelled as "sentence" in the "Level" column in the table) were computed at the sentence level, then averaged for the treebank. Treebank-level metrics (labelled as "treebank" in the "Level" column in the table), were calculated regardless of sentence boundaries.

A balance between morphological metrics (e.g. number of word forms per lemma) and syntactic metrics (e.g. number of clauses per sentence) was one of the objectives of this selection. A number of primitive lexical complexity metrics (e.g. type token ratio, token count) were also included. In this sense, noun to verb ratio is commonly used in readability research, as a preponderance of verbs signals more complex sentence. A similar case can be made for the number of clauses per sentence, as a sentence with many subordinate clauses will be naturally harder to process than a singular main clause.

We discuss also a pair of metrics not commonly seen in research and only enabled by the presence of dependency annotation, namely number of \emph{ccomp} and \emph{xcomp} relations in a sentence respectively. The intuition behind these metrics is that processing \emph{ccomp} as opposed to \emph{xcomp} relations would entail a lower cognitive load. Thus, a higher usage of \emph{xcomp} might suggest higher overall complexity. Length of longest dependecy link is likewise enabled by the nature of the annotation, and gives an idea of how close related elements are on the surface level - academic Hungarian, as an example, tends do displace items across long sequences of clauses, often separating elements of the main clause by multiple lines of text.

Although there are more sophisticated metrics on the market, this set was ultimately selected for ease of calculation and because their interpretation is generally approachable without further analytics, which woudln't necesserily be the case for more nuanced strategies.

\section{Results}

\subsection{Official treebanks}

\input{tables/results.tex}

Table~\ref{tab:results} shows the values related to each metric and each treebank, with the top 3 values highlighted for each metric. For the \emph{cmn\_pud} treebank, \emph{wfpl} is not reported, as this treebank does not include lemma information.

Similary to \citep{berdicevskis-etal-2018-using}, multiple treebanks were used for each language to get a sense of metric robustness, i.e. whether the metric returns consistent results for different data in the same language. Most metrics are well-behaved in this sense, with \emph{en\_atis} being the biggest outlier across the board. The values from this treebank differ substantially from \emph{en\_gum} and \emph{en\_ewt} in most metrics. One possible justification for this outcome is the nature of how the data was sourced for \emph{en\_atis}, namely by collecting and transcribing user interactions with automated inquiry systems related to flight information.

Outside of this case, most metrics yield comparable outputs for most language treebank sets, a somewhat unexpected outcome in reference to the forecasted effect of at times wildly different treebank sizes. The biggest difference was measured between the two German treebanks, \emph{de\_gsd} and \emph{de\_hdt}, with the latter being over five times larger than the former. Despite this difference, outputs for the two treebanks are similar for all metrics except \emph{wfpl} and \emph{vps}, which were already projected to suffer most from the effect of treebank size (more sentences represent more opportunities for rarer POS tag sequences and lemma forms to appear).

For the robustness aspect, it should be noted that the two treebanks for Hungarian score very similarly across the board, however this is a somewhat unremarkable outcome. The two treebanks were in fact derived from splitting a single original one in half, therefore they are smaller than all other treebanks in the experiment, and, unlike those other treebanks, have the same source. Conclusions derived from this observation are therefore necessarily weaker than those drawn from, for example, the two German treebanks.

When it comes to the specific outputs generated during the experiment, it is generally the case that Hungarian and Chinese treebanks yield values that suggest a higher degree of overall complexity. All metrics except \emph{ttr} (type token ratio), \emph{wfpl} (word forms per lemma) and \emph{vps} (variability of 3 POS tag sequences) show at least either language in the top ranks. Notably, two of these metrics - \emph{wfpl} and \emph{vps} - were projected to be most affected by treebank size, and Hungarian and Chinese have the smallest treebanks in this project's dataset. It should also not go without mention that there is a relation between \emph{tc} and \emph{ttr}: \emph{tc} values are highest for exactly the 4 Hungarian and Chinese treebanks, thus partially explaining the lower expression of \emph{ttr}.

For the metric "ratio of verbs with explicit subject", two variants were reported - \emph{ves} and \emph{ves\_t}, the former computed at the sentence level and the latter at the treebank level. This decision was motivated by the presence of several nominal sentences in the treebanks, which made this metric unstable at the sentence level. Seeing the results, even though the numbers and rankings obtained differ, German and Hungarian are still the top languages. The fact that English (which usually requires verbal subjects to be present) has fewer explicit subjects on average than Hungarian (which doesn't) is an unexpected outcome. However, this might be caused by discrepancies in annotation conventions, and not necessarily a true difference between the languages.

The metrics \emph{ccomp} (average number of ccomp relations per sentence) and \emph{xcomp} (average number of xcomp relations per sentence) were a somewhat novel selection in this experiment. An argument was made above that a usage of the xcomp dependency relation might suggest higher overall complexity. Althugh for both \emph{ccomp} and \emph{xcomp} English and Chinese were the highest values, the \emph{xcomp} metric is higher than \emph{ccomp} in all treebanks except \emph{cmn\_gsd}. In this sense \emph{cmn\_gsd} is the only instance of higher ccomp relation usage, suggesting Chinese to be the least complex language according to the presuppositions related to these metrics - but this case cannot be confidently made without more detailed analysis of how these relations function in the specific treebanks.

\subsection{Parallel treebank}

\input{tables/results_ours.tex}

To corroborate the observations drawn from the offical UD treebanks, the same set of metrics were computed on the parallel corpus introduced above. Table~\ref{tab:results-ours} outlines the results of this second part.

Hungarian and Chinese remain the holders of the highest values in 8 out of 12 (11 if we group \emph{ves} and \emph{ves\_t}), lending credence to the theory that these languages are more complex in relation to these metrics.

The metrics \emph{vps} and \emph{wfpl} were highlighted as affected by treebank size in the section above, therefore conferming them on same-size treebanks, however small, might still be a valuable undertaking. Indeed, Hungarian, a language with complex morphology, takes the lead in \emph{wfpl} and Chinese outranks German and English, which have a more constrained word order.

Generally, while the results on the parallel treebank are more or less in line with the official treebanks, the sample size in this case is too small. Thus, meaningful conclusions cannot be drawn. Nevertheless the goal was to highlight the usage of parallel treebanks as a strategy to control for as many factors as possible and isolate aspects related to complexity.

\section{Discussion and future research}

This project attempted to showcase the use of treebanks in the UD framework for complexity research. To achieve this, a set of treebanks across four languages and a set of text-level metrics related to complexity were selected. The previous sections detailed the results of the experiment.

As mentioned before, not all treebanks are created equal. In the case of this experiment, it was both the case that the treebank had different sizes, and that it is not a guarantee that annotation conventions across language frameworks are harmonic by default. \citep{berdicevskis-etal-2018-using}, a study similar to this work, took into account this latter obstacle by smoothing the inter-language annotation differences, while the current work took no such approaches. An idea for future work on the topic might be to look at [AUX + VERB + (N/C)SUBJ] structures in relation to the \emph{ves} metric, as the subject can be confusingly attached to either VERB or AUX.

\citep{berdicevskis-etal-2018-using} employed a variety of nuanced metrics, which were certainly better suited to the task of gauging language complexity. Although the selection of metrics here can be defended in terms of interpretability and low hardware requirements, future work on the topic should err toward the side of significance rather than interpretability.

On the topic of significance, statistical tests are absent from this experiment, although the results obtained would have benefitted from them. Breaking academic formality for this sentence, I will do my best to learn how to perform them for my future projects.

While no claims are made with regard to overall language complexity or even treebank complexity, perhaps this work eases the way to future experimentation with UD treebanks and the general field of language complexity.

\section*{Acknowledgements}

I thank Dr. Çağrı Çöltekin for his patient counselling and for providing insight into his 2018 paper, which served as inspiration for most of this work.

\bibliography{anthology,cited}

\appendix

\section*{Resources}

\label{sec:appendix}

All code for this paper is available at \url{https://github.com/waron97/dtdp-final-paper}.

\end{document}

### 2023-03-27 paper/paper.tex
update paper

@@
 
-\title{Cross-linguistic complexity analysis using UD treebanks}
+\title{Examining readability of text generated through GPT-3 }
 
@@
   University of Tübungen \\
-  \texttt{email@domain}}
+  \texttt{winkler.aron5@gmail.com}}
 
@@
 \begin{abstract}
-    The aim of this project is to explore the use of UD treebanks in cross-linguistic complexity research.
-While it is not this paper's objective to make strong theoretical claims on complexity in its own right, it examines complexity measures computed on various UD treebanks across 4 languages.
-This work focuses on treebanks for English, German, Hungarian, and Chinese, in an attempt to determine whether complexity measures stay consistent within treebanks for the same language, and whether some languages consistently show higher complexity than others.
-    These claims are verified with official UD corpora, as well as with a parallel corpus of 50 sentences for each language mentioned above.
+
 \end{abstract}
@@
 \section{Introduction}
-
-What constitutes linguistic complexity is an elusive multi-faceted inquiry, and quantifying it is equally challenging.
-Previous approaches have included text-based metrics as well as eye-tracking studies (e.g.
-\citealp{Lee:2007}), but a general consensus has not been reached, especially when it comes to the notion of "overall complexity" of a language.
-In the case that an experiment produces numeric values as a proxy for complexity, it is not obvious that results computed on two different languages are comparable.
-
-In this work, the objective is not to attempt to resolve these foundational question about linguistic complexity, but rather to accept that there are certain well documented text-based proxies for computing it - such as lexical, morphological, and syntactic complexity measures - and to explore how the Universal Dependencies framework can be employed to further the subject matter.
-
-Indeed, the potential of UD-annotated data appears to be largely untapped in linguistic complexity research.
-A convincing contribution to this field was made by \citealp{berdicevskis-etal-2018-using}, estimating robustness of complexity values computed on a variety of treebanks, while accounting for additional obstacles such as language-specific annotation conventions that might reduce comparability.
-
-While this project is more restricted in scope and claims, the goal is to showcase the use of UD treebanks in linguistic complexity research along the same lines as \citealp{berdicevskis-etal-2018-using}.
-For this purpose, a variety of treebanks across four languages were used, in combination with a set of text-based complexity metrics inspired in part by readability research, and in part by complexity research more broadly.
-These metrics are computed on two datasets, one of which is a collection of official UD treebanks, and the other being a small in-house parallel corpus comprising 50 sentences of each of the 4 target languages.
-
+\citep{corenlp} hello
 \section{Dataset and metrics}
@@
 
-\input{tables/webdata.tex}
-
-Table~\ref{tab:webdata} shows the list of UD web treebanks used in this experiment.
-For the English language, which has many treebanks available to it, the Atis treebank (Github\footnote{\url{https://github.com/UniversalDependencies/UD_English-Atis}}), the gum GUM (\citealp{berzak2016tle}, GitHub\footnote{\url{https://github.com/UniversalDependencies/UD_English-GUM}}), and the EWT treebank (\citealp{silveira14gold}, GitHub\footnote{\url{https://github.com/UniversalDependencies/UD_English-EWT}}) were ised.
-For German, this work employs GSD (\citealp{mcdonald-etal-2013-universal}, GitHub \footnote{\url{https://github.com/UniversalDependencies/UD_German-GSD}}) and HDT (\citealp{borges-volker-etal-2019-hdt}, \citealp{hennig-kohn-2017-dependency}, \citealp{hdtextra}, \citealp{hdtguide}, GitHub \footnote{\url{https://github.com/UniversalDependencies/UD_German-HDT}}).
-For Chinese,
-GSD (GitHub\footnote{\url{https://github.com/UniversalDependencies/UD_Chinese-GSD}}) and PUD (GitHub\footnote{\url{https://github.com/UniversalDependencies/UD_Chinese-PUD}}) were the final choice.
-For Hungarian, only the Szeged treebank (\citealp{szeged}, Github\footnote{\url{https://github.com/UniversalDependencies/UD_Hungarian-Szeged}}) was available - therefore in order to have at least two treebanks for each language, the sentences annotated in the Szeged treebank were randomly distributed across two new treebanks (hun\_szeged\_1, hun\_szeged\_2).
-
-An additional parallel treebank\footnote{\url{https://github.com/iscl-dtdp/ParallelTreebank-FinalProject}} made by Nino Meisinger, Qin Gu, Lisa Wang, and Aron Winkler as coursework at the University of Tübingen was also employed.
-Although its limited size of only 50 sentences per language prevents meaningful conclusions from being drawn from measurements computed on it, its parallel nature nevertheless allows interesting observations to be made in reference to the complexity metric values obtained from the official treebanks.
-Sentences for this parallel treebank were sourced from Tatoeba\footnote{\url{https://tatoeba.org/en/}}, more specifically by selecting the 50 longest sentences that had translations for all 4 target languages.
-If more than one translation was available for any given language, then the choice was left up to the annotator.
-The final treebank was then manually annotated according to UD standards by annotators native or very proficient in the target language.
-
 \subsection{Metrics}
-
-\input{tables/metrics.tex}
-
-Table~\ref{tab:metrics} details the complexity metrics used in this experiment.
-Most of them were inspired by readability research, while a couple were adopted from \citep{berdicevskis-etal-2018-using}.
-Sentence-level metrics (labelled as "sentence" in the "Level" column in the table) were computed at the sentence level, then averaged for the treebank.
-Treebank-level metrics (labelled as "treebank" in the "Level" column in the table), were calculated regardless of sentence boundaries.
-
-A balance between morphological metrics (e.g.
-number of word forms per lemma) and syntactic metrics (e.g.
-number of clauses per sentence) was one of the objectives of this selection.
-A number of primitive lexical complexity metrics (e.g.
-type token ratio, token count) were also included.
-In this sense, noun to verb ratio is commonly used in readability research, as a preponderance of verbs signals more complex sentence.
-A similar case can be made for the number of clauses per sentence, as a sentence with many subordinate clauses will be naturally harder to process than a singular main clause.
-
-We discuss also a pair of metrics not commonly seen in research and only enabled by the presence of dependency annotation, namely number of \emph{ccomp} and \emph{xcomp} relations in a sentence respectively.
-The intuition behind these metrics is that processing \emph{ccomp} as opposed to \emph{xcomp} relations would entail a lower cognitive load.
-Thus, a higher usage of \emph{xcomp} might suggest higher overall complexity.
-Length of longest dependecy link is likewise enabled by the nature of the annotation, and gives an idea of how close related elements are on the surface level - academic Hungarian, as an example, tends do displace items across long sequences of clauses, often separating elements of the main clause by multiple lines of text.
-
-Although there are more sophisticated metrics on the market, this set was ultimately selected for ease of calculation and because their interpretation is generally approachable without further analytics, which woudln't necesserily be the case for more nuanced strategies.
 
@@
 
-\subsection{Official treebanks}
-
-\input{tables/results.tex}
-
-Table~\ref{tab:results} shows the values related to each metric and each treebank, with the top 3 values highlighted for each metric.
-For the \emph{cmn\_pud} treebank, \emph{wfpl} is not reported, as this treebank does not include lemma information.
-
-Similary to \citep{berdicevskis-etal-2018-using}, multiple treebanks were used for each language to get a sense of metric robustness, i.e.
-whether the metric returns consistent results for different data in the same language.
-Most metrics are well-behaved in this sense, with \emph{en\_atis} being the biggest outlier across the board.
-The values from this treebank differ substantially from \emph{en\_gum} and \emph{en\_ewt} in most metrics.
-One possible justification for this outcome is the nature of how the data was sourced for \emph{en\_atis}, namely by collecting and transcribing user interactions with automated inquiry systems related to flight information.
-
-Outside of this case, most metrics yield comparable outputs for most language treebank sets, a somewhat unexpected outcome in reference to the forecasted effect of at times wildly different treebank sizes.
-The biggest difference was measured between the two German treebanks, \emph{de\_gsd} and \emph{de\_hdt}, with the latter being over five times larger than the former.
-Despite this difference, outputs for the two treebanks are similar for all metrics except \emph{wfpl} and \emph{vps}, which were already projected to suffer most from the effect of treebank size (more sentences represent more opportunities for rarer POS tag sequences and lemma forms to appear).
-
-For the robustness aspect, it should be noted that the two treebanks for Hungarian score very similarly across the board, however this is a somewhat unremarkable outcome.
-The two treebanks were in fact derived from splitting a single original one in half, therefore they are smaller than all other treebanks in the experiment, and, unlike those other treebanks, have the same source.
-Conclusions derived from this observation are therefore necessarily weaker than those drawn from, for example, the two German treebanks.
-
-When it comes to the specific outputs generated during the experiment, it is generally the case that Hungarian and Chinese treebanks yield values that suggest a higher degree of overall complexity.
-All metrics except \emph{ttr} (type token ratio), \emph{wfpl} (word forms per lemma) and \emph{vps} (variability of 3 POS tag sequences) show at least either language in the top ranks.
-Notably, two of these metrics - \emph{wfpl} and \emph{vps} - were projected to be most affected by treebank size, and Hungarian and Chinese have the smallest treebanks in this project's dataset.
-It should also not go without mention that there is a relation between \emph{tc} and \emph{ttr}: \emph{tc} values are highest for exactly the 4 Hungarian and Chinese treebanks, thus partially explaining the lower expression of \emph{ttr}.
-
-For the metric "ratio of verbs with explicit subject", two variants were reported - \emph{ves} and \emph{ves\_t}, the former computed at the sentence level and the latter at the treebank level.
-This decision was motivated by the presence of several nominal sentences in the treebanks, which made this metric unstable at the sentence level.
-Seeing the results, even though the numbers and rankings obtained differ, German and Hungarian are still the top languages.
-The fact that English (which usually requires verbal subjects to be present) has fewer explicit subjects on average than Hungarian (which doesn't) is an unexpected outcome.
-However, this might be caused by discrepancies in annotation conventions, and not necessarily a true difference between the languages.
-
-The metrics \emph{ccomp} (average number of ccomp relations per sentence) and \emph{xcomp} (average number of xcomp relations per sentence) were a somewhat novel selection in this experiment.
-An argument was made above that a usage of the xcomp dependency relation might suggest higher overall complexity.
-Althugh for both \emph{ccomp} and \emph{xcomp} English and Chinese were the highest values, the \emph{xcomp} metric is higher than \emph{ccomp} in all treebanks except \emph{cmn\_gsd}.
-In this sense \emph{cmn\_gsd} is the only instance of higher ccomp relation usage, suggesting Chinese to be the least complex language according to the presuppositions related to these metrics - but this case cannot be confidently made without more detailed analysis of how these relations function in the specific treebanks.
-
-\subsection{Parallel treebank}
-
-\input{tables/results_ours.tex}
-
-To corroborate the observations drawn from the offical UD treebanks, the same set of metrics were computed on the parallel corpus introduced above.
-Table~\ref{tab:results-ours} outlines the results of this second part.
-
-Hungarian and Chinese remain the holders of the highest values in 8 out of 12 (11 if we group \emph{ves} and \emph{ves\_t}), lending credence to the theory that these languages are more complex in relation to these metrics.
-
-The metrics \emph{vps} and \emph{wfpl} were highlighted as affected by treebank size in the section above, therefore conferming them on same-size treebanks, however small, might still be a valuable undertaking.
-Indeed, Hungarian, a language with complex morphology, takes the lead in \emph{wfpl} and Chinese outranks German and English, which have a more constrained word order.
-
-Generally, while the results on the parallel treebank are more or less in line with the official treebanks, the sample size in this case is too small.
-Thus, meaningful conclusions cannot be drawn.
-Nevertheless the goal was to highlight the usage of parallel treebanks as a strategy to control for as many factors as possible and isolate aspects related to complexity.
-
 \section{Discussion and future research}
-
-This project attempted to showcase the use of treebanks in the UD framework for complexity research.
-To achieve this, a set of treebanks across four languages and a set of text-level metrics related to complexity were selected.
-The previous sections detailed the results of the experiment.
-
-As mentioned before, not all treebanks are created equal.
-In the case of this experiment, it was both the case that the treebank had different sizes, and that it is not a guarantee that annotation conventions across language frameworks are harmonic by default.
-\citep{berdicevskis-etal-2018-using}, a study similar to this work, took into account this latter obstacle by smoothing the inter-language annotation differences, while the current work took no such approaches.
-An idea for future work on the topic might be to look at [AUX + VERB + (N/C)SUBJ] structures in relation to the \emph{ves} metric, as the subject can be confusingly attached to either VERB or AUX.
-
-\citep{berdicevskis-etal-2018-using} employed a variety of nuanced metrics, which were certainly better suited to the task of gauging language complexity.
-Although the selection of metrics here can be defended in terms of interpretability and low hardware requirements, future work on the topic should err toward the side of significance rather than interpretability.
-
-On the topic of significance, statistical tests are absent from this experiment, although the results obtained would have benefitted from them.
-Breaking academic formality for this sentence, I will do my best to learn how to perform them for my future projects.
-
-While no claims are made with regard to overall language complexity or even treebank complexity, perhaps this work eases the way to future experimentation with UD treebanks and the general field of language complexity.
 
@@
 
-I thank Dr. Çağrı Çöltekin for his patient counselling and for providing insight into his 2018 paper, which served as inspiration for most of this work.
-
-\bibliography{anthology,cited}
+\bibliography{anthology,custom}
 
 \appendix
-
 \section*{Resources}
@@
 
-All code for this paper is available at \url{https://github.com/waron97/dtdp-final-paper}.
+All code for this paper is available at \url{https://github.com/waron97/nlp_for_readability_term_paper}.
 

### 2023-03-27 paper/paper.tex
generate plots and update paper

@@
 \documentclass[11pt]{article}
+
+\usepackage{graphicx}
 
@@
 \begin{abstract}
-
+    This paper examines readability-related metrics of AI-generated text, specifically texts produced by OpenAI's GPT-3.
+Generation was performed across 10 topics and the 6 proficiency levels defined by the CEFR framework, and outputs were compared with Simple Wikipedia and (standard) Wikipedia articles on the same topics.
 \end{abstract}
@@
 \section{Introduction}
-\citep{corenlp} hello
+
+AI text generation has experiences a massive rise in popularity with the media reach achieved by the release of OpenAI's GPT-3 \citep{brown2020language} and its chat GUI interface ChatGPT\footnote{\url{https://openai.com/blog/chatgpt}}.
+While early public opinion focused on how - in the school setting - it would be students who would benefit from the new system by having their schoolwork done for them, little was said about how AI text generation could become an asset for didactitians in generating teaching material.
+
+AI text generation might become an important tool for teachers, allowing them to generate learning material about topics and events that engage their students that might not be available on the market.
+An interesting setting for such a phenomenon is language learning, where it is necessary to produce learning material for students at different proficiency levels: whether to create ex-novo or simplify existing text material, AI text-generation tools can easily be imagined as having an enabling effect.
+
+It is however not obvious that AI generation tools reach the desired quality or adherence to the didactic goals that teachers set out.
+Owing to the increased popularity of these tools, more research is desirable on their applicability in a number of different settings.
+
+This work aims to make a contribution to those questions by verifying the readability level of text content generated by GPT-3 with regard to 10 topics and the 6 proficiency levels defined by the CEFR framework.
+A number of complexity metrics are computed on the generated texts and compared with Simple Wikipedia\footnote{\url{https://simple.wikipedia.org/wiki/Main_Page}} as well as Wikipedia\footnote{\url{https://www.wikipedia.org/}} articles on the same topics in order to gauge the true readability levels of the model outputs.
 \section{Dataset and metrics}
+\subsection{Dataset}
 
-\subsection{Dataset}
+OpenAI's REST api was employed to produce texts with the "text-davinci-003" model (a GPT-3 variant) across 10 topics\footnote{'Color Blindness, The Great Depression, Butterflies, Dogs, Semantics, The Internet, The Moon, Dinosaurs, Economics, Quantum Mechanics'} and 6 CEFR proficiency levels\footnote{A1, A2, B1, B2, C1, C2}.
+Topics were selected without any particular guiding principle other than a general goal of variety between events, entities, and phenomena.
+The prompts fed to the model followed the template:
+
+\begin{quote}
+  \emph{Write a text for learners of English at the [level] level about [topic].}
+\end{quote}
+
+In addition to the 60 texts obtained in this way, the dataset was enriched by Simple Wikipedia and Wikipedia articles corresponding to each of the 10 topics, resulting in an additional 20 dataset entries.
+
+\subsection{Preprocessing}
+
+Preprocessing of the dataset comprised enrichment with POS and dependency annotation.
+Part-of-speech tagging was achieved with nltk \citep{LoperBird02} and dependency relations were added through nltk in conjunction with CoreNLP models \citep{corenlp}.
 
@@
 
+\input{tables/metrics.tex}
+
+Table~\ref{tab:metrics} lists the readability metrics used in this experiment.
+This specific selection was inspired by \citep{venturi-etal-2015-nlp}, but many more examples of their usage can be found.
+
+This feature set almost exclusively targets lexis and syntax, with little regard for morphology outside of the noun-to-verb ratio feature.
+While this is a limitation of this set, the language of this experiment is solely English, a language not necessarily known for a complex morphological system.
+
+The feautre "Words in Top 3000 English words" uses unigram frequencies from the Google Web Trillion Word corpus\footnote{\url{https://ai.googleblog.com/2006/08/all-our-n-gram-are-belong-to-you.html}} to isolate the 3000 most frequent words of the English language.
+Many research approaches to readability employ similar ideas to gain insight into the lexical complexity of texts, such as the BIV (basic Italian vocabulary) employed by \citep{dellorletta-etal-2011-read}.
+
+All feature values outside of type-token-ratio (which ignores sentence boundaries) are computed at the sentence level.
+When a dataset entry comprises multiple sentences, the output for that entry is the average of the values for the sentences that compose it.
+
 \section{Results}
+
+\input{tables/means.tex}
+
+Table~\ref{tab:means} details the average output for each difficulty level and metric. For each metric, the 3 values most indicative of low readability are highlighted.
+
+With regards to the selected metrics, GPT-3 does a good job generating texts for the target readability level in the sense that prompts requesting C1 and C2 level material on average yield the least readable texts, whereas lower proficiency levels tend to have scores indicating higher readability.
+
+While no rigorous checks are carried out in this paper in reference to the correlation between these metrics and article readability levels, it's still possible to examine some of the results to gain some insight into the black box that is GPT-3.
+
+\input{figs/wit_means.tex}
+
+Among the selected measures, WIT (ratio of words in the 3000 most frequent English words) appears to be the most well-behaved, insofar as the prompted CEFR levels mostly maintain their natural progression.
+Figure~\ref{fig:wit-mean} shows the isolated scores for the WIT metric, which show a correlation with difficulty.
+Wikipedia and Simple Wikipedia articles have lower values, however this is possibly connected to those texts being several times longer.
 

### 2023-03-27 main.py

@@ -2,7 +2,7 @@
 from src.simple_wiki import load_simple_wiki_sql
 from dotenv import load_dotenv
 from src.make_dataset import make_dataset, write_dataset
-from src.experiment import analyze_dataset
+from src.experiment import analyze_dataset, plot_results
 
 load_dotenv()
 
@@ -11,7 +11,8 @@ def main():
     # load_simple_wiki_sql()
     # make_dataset()
     # write_dataset()
-    analyze_dataset()
+    metrics, experiment = analyze_dataset()
+    plot_results(metrics, experiment)
 
 
 if __name__ == "__main__":

### 2023-03-27 src/constants/core.py

@@ -1,0 +1,34 @@
+TOPICS = [
+    "Color Blindness",
+    "The Great Depression",
+    "Butterflies",
+    "Dogs",
+    "Semantics",
+    "The Internet",
+    "The Moon",
+    "Dinosaurs",
+    "Economics",
+    "Quantum Mechanics",
+]
+
+LEVELS = [
+    "A1",
+    "A2",
+    "B1",
+    "B2",
+    "C1",
+    "C2"
+]
+
+WIKI_TOPICS = [
+    "Color Blindness",
+    "Great Depression",
+    "Butterfly",
+    "Dog",
+    "Semantics",
+    "Internet",
+    "Moon",
+    "Dinosaur",
+    "Economics",
+    "Quantum Mechanics",
+]

### 2023-03-27 src/experiment/__init__.py

@@ -1,1 +1,2 @@
 from .main import analyze_dataset
+from .plot_results import plot_results

### 2023-03-27 src/experiment/entry.py

@@ -5,6 +5,8 @@ from ..type import Dependencies
 
 class Entry(TypedDict):
     id: str
+    level: str
+    topic: str
     text: str
     pos: POS_TAGS
     parse: List[Dependencies]

### 2023-03-27 src/experiment/experiment.py

@@ -6,18 +6,22 @@ from .metrics import Metric
 from ..type import Dependencies
 from tqdm import tqdm
 import pickle
+import re
 
 
 class Dataset:
     def __init__(self) -> None:
         self.data: List[Entry] = []
 
-    def add_sent(self, id: str, text: str):
+    def add_sent(self, id: str, text: str, level: str, topic: str):
         self.data.append({
             "id": id,
             "text": text,
             "pos": [],
-            "parse": []
+            "parse": [],
+            "level": level,
+            "topic": topic,
+            "tree_height": 0
         })
 
 
@@ -25,6 +29,8 @@ class Result(TypedDict):
     id: str
     metric: str
     value: float
+    level: str
+    topic: str
 
 
 class ResultSet:
@@ -40,12 +46,26 @@ class ResultSet:
     def all(self):
         return self.results
 
-    def filter(self, ids: List[str], metrics: List[str]) -> List[Result]:
-        results: List[Result] = []
-        for result in self.results:
-            if result["id"] in ids or result["metric"] in metrics:
-                results.append(result)
-        return results
+    def filter(self, ids: List[str] = [], metrics: List[str] = [], idlike: str = "", topics: List[str] = [], levels: List[str] = []):
+        results: List[Result] = self.all()
+        if ids:
+            results = [result for result in results if result["id"] in ids]
+        if metrics:
+            results = [
+                result for result in results if result["metric"] in metrics]
+        if topics:
+            results = [
+                result for result in results if result["topic"] in topics]
+        if levels:
+            results = [
+                result for result in results if result["level"] in levels]
+        if idlike:
+            results = [result for result in results if idlike in result["id"]]
+        return ResultSet(results)
+
+    def mean(self, metric: str) -> float:
+        results = self.filter(metrics=[metric]).all()
+        return sum([result["value"] for result in results]) / len(results)
 
 
 class Experiment:
@@ -80,7 +100,10 @@ class Experiment:
                 results.append({
                     "id": entry["id"],
                     "metric": metric.name,
-                    "value": value})
+                    "value": value,
+                    "level": entry["level"],
+                    "topic": entry["topic"]
+                })
         self.results = ResultSet(results)
 
     def load_dataset_from_pkl(self, fpath: str):

### 2023-03-27 src/experiment/main.py

@@ -13,14 +13,14 @@ from .metrics import (AvgSentenceLength, ClausesPerSentence,
 def _make_dataset(testset: TestSet) -> Dataset:
     dataset = Dataset()
     for entry in testset["gpt_generation"]:
-        _id = f"{entry['topic']}-{entry['level']}"
-        dataset.add_sent(_id, entry["text"])
+        _id = f"{entry['level']}-{entry['topic']}"
+        dataset.add_sent(_id, entry["text"], entry["level"], entry["topic"])
     for entry in testset["wikipedia"]["simple"]:
         _id = f"wikisimple-{entry['topic']}"
-        dataset.add_sent(_id, entry["text"])
+        dataset.add_sent(_id, entry["text"], "wikisimple", entry["topic"])
     for entry in testset["wikipedia"]["standard"]:
         _id = f"wikistandard-{entry['topic']}"
-        dataset.add_sent(_id, entry["text"])
+        dataset.add_sent(_id, entry["text"], "wikistandard", entry["topic"])
     return dataset
 
 
@@ -55,3 +55,5 @@ def analyze_dataset():
 
     experiment.compute_metrics()
     dump_to_file(experiment.results.all(), "data/results.json")
+
+    return metrics, experiment

### 2023-03-27 src/experiment/plot_results.py

@@ -1,0 +1,107 @@
+from typing import List
+
+from ..experiment.metrics import Metric
+from ..experiment.experiment import Experiment
+from matplotlib import pyplot as plt
+from ..constants import TOPICS, LEVELS
+import pandas as pd
+
+
+def _map_metric_name(name):
+    if name == "wikisimple":
+        return "S Wiki"
+    elif name == "wikistandard":
+        return "Wiki"
+    return name
+
+
+def _make_table(metrics: List[Metric], experiment: Experiment):
+    columns = [metric.abbreviation for metric in metrics]
+    index = LEVELS + ["wikisimple", "wikistandard"]
+    index_mapped = list(map(_map_metric_name, index))
+
+    df = pd.DataFrame(index=index_mapped, columns=columns)
+    for metric in metrics:
+        for level in index:
+            try:
+                value = experiment.results.filter(
+                    metrics=[metric.name], levels=[level]).mean(metric.name)
+                value = round(value, 3)
+                df.loc[_map_metric_name(level), metric.abbreviation] = value
+            except ZeroDivisionError:
+                print(
+                    f"ZeroDivisionError for {metric.abbreviation} and {level}! Skipping...")
+    print(df.to_latex())
+
+
+def _make_topic_metric_plot(metric: str, topic: str, experiment: Experiment, short_name: str = ""):
+    values = experiment.results.filter(
+        metrics=[metric], idlike=topic).all()
+
+    levels = [_map_metric_name(value["level"]) for value in values]
+    tups = [(level, value["value"]) for level, value in zip(levels, values)]
+    tups.sort(key=lambda tup: tup[1])
+
+    x = list(range(len(tups)))
+    height = [tup[1] for tup in tups]
+    labels = [tup[0] for tup in tups]
+    plt.bar(x, height, tick_label=labels, width=0.8,)
+    plt.xticks(rotation=45)
+    fname = f"plot_{short_name}_{topic}.png" if short_name else f"plot_{metric}_{topic}.png"
+    plt.savefig(f"data/analysis/topics/{fname}")
+    plt.cla()
+
+
+def _make_plot(metric: str, experiment: Experiment, short_name: str = "", sort_revert: bool = False):
+    a1 = experiment.results.filter(
+        metrics=[metric], levels=["A1"]).mean(metric)
+    a2 = experiment.results.filter(
[…]

### 2023-03-27 src/make_dataset/make_dataset.py

@@ -9,42 +9,7 @@ from sqlalchemy.orm import Session
 from typing import List, Tuple
 from tqdm import tqdm
 from ..util import dump_to_file
-
-
-TOPICS = [
-    "Color Blindness",
-    "The Great Depression",
-    "Butterflies",
-    "Dogs",
-    "Semantics",
-    "The Internet",
-    "The Moon",
-    "Dinosaurs",
-    "Economics",
-    "Quantum Mechanics",
-]
-
-LEVELS = [
-    "A1",
-    "A2",
-    "B1",
-    "B2",
-    "C1",
-    "C2"
-]
-
-WIKI_TOPICS = [
-    "Color Blindness",
-    "Great Depression",
-    "Butterfly",
-    "Dog",
-    "Semantics",
-    "Internet",
-    "Moon",
-    "Dinosaur",
-    "Economics",
-    "Quantum Mechanics",
-]
+from ..constants import TOPICS, LEVELS, WIKI_TOPICS
 
 
 def _get_standard_wiki_text(topic):

### 2023-03-27 wiki-clean/index.js
remove console log

@@ -32,7 +32,6 @@ app.post("/clean", (req, res, next) => {
 app.post("/article", (req, res, next) => {
   try {
     const { title } = req.body;
-    console.log("title", title);
     return getNormalArticle(title)
       .then((text) => {
         res.json({ text });

### 2023-03-27 main.py
add script to create classifier dataset

@@ -1,5 +1,5 @@
 
-from src.simple_wiki import load_simple_wiki_sql
+from src.simple_wiki import load_simple_wiki_sql, load_full_wikipedia_dataset_sql
 from dotenv import load_dotenv
 from src.make_dataset import make_dataset, write_dataset
 from src.experiment import analyze_dataset, plot_results
@@ -8,6 +8,7 @@ load_dotenv()
 
 
 def main():
+    load_full_wikipedia_dataset_sql()
     # load_simple_wiki_sql()
     # make_dataset()
     # write_dataset()

### 2023-03-27 src/simple_wiki/__init__.py

@@ -1,1 +1,1 @@
-from .to_sql import load_simple_wiki_sql
+from .to_sql import load_simple_wiki_sql, load_full_wikipedia_dataset_sql

### 2023-03-27 src/simple_wiki/to_sql.py

@@ -3,15 +3,15 @@ from xml.etree import ElementTree as ET
 
 import requests
 from ..sql import engine
-from ..sql.models import SimpleWikiEntry
+from ..sql.models import SimpleWikiEntry, WikiArticle
 from tqdm import tqdm
 from sqlalchemy.orm import Session
 from ..util import dump_to_file
 
 
-def _exists(id):
+def _exists(id, model=SimpleWikiEntry):
     with Session(engine) as session:
-        return session.query(SimpleWikiEntry).filter_by(id=id).first() is not None
+        return session.query(model).filter_by(id=id).first() is not None
 
 
 def _save_wiki(id, title, text, text_clean, session: Session = None):
@@ -19,6 +19,27 @@ def _save_wiki(id, title, text, text_clean, session: Session = None):
         return
     entry = SimpleWikiEntry(
         id=id, title=title, text=text, text_clean=text_clean)
+    session.add(entry)
+
+
+def get_standard_article(title):
+    try:
+        content = requests.post(
+            "http://localhost:6000/article",
+            headers={"Content-Type": "application/json"},
+            data=json.dumps({"title": title})
+        ).content
+        l = json.loads(content)
+        return l["text"]
+    except:
+        return ""
+
+
+def _save_article(id, title, standard, simple, session: Session = None):
+    if not title or not standard or not simple:
+        return
+    entry = WikiArticle(
+        id=id, title=title, standard_text=standard, simple_text=simple)
     session.add(entry)
 
 
@@ -52,4 +73,26 @@ def load_simple_wiki_sql():
             text_clean = get_clean_text(text)
             if not _exists(id):
                 _save_wiki(id, title, text, text_clean, session=session)
+        session.commit()
+
+
+def load_full_wikipedia_dataset_sql():
+    p = "data/simple_wiki/page-articles.xml"
+    page_tag = "{http://www.mediawiki.org/xml/export-0.10/}page"
+    id_tag = "{http://www.mediawiki.org/xml/export-0.10/}id"
+    title_tag = "{http://www.mediawiki.org/xml/export-0.10/}title"
+    text_tag = "{http://www.mediawiki.org/xml/export-0.10/}revision/{http://www.mediawiki.org/xml/export-0.10/}text"
+    tree = ET.parse(p)
+    root = tree.getroot()
+    pages = root.findall(page_tag)
+    with Session(engine) as session:
+        for page in tqdm(pages):
+            id = page.find(id_tag).text
+            title = page.find(title_tag).text
[…]

### 2023-03-27 src/sql/models.py

@@ -46,6 +46,15 @@ class ClearCorpusEntry(Base):
     cml2ri: Mapped[Optional[float]]
 
 
+class WikiArticle(Base):
+    __tablename__ = "wiki_article"
+
+    id: Mapped[int] = mapped_column(primary_key=True)
+    title: Mapped[str]
+    standard_text: Mapped[str]
+    simple_text: Mapped[str]
+
+
 class DatasetEntry(Base):
     __tablename__ = "dataset"
 

### 2023-03-28 src/simple_wiki/to_sql.py
update fn

@@ -9,7 +9,9 @@ from sqlalchemy.orm import Session
 from ..util import dump_to_file
 
 
-def _exists(id, model=SimpleWikiEntry):
+def _exists(id, model=SimpleWikiEntry, session: Session = None):
+    if session:
+        return session.query(model).filter_by(id=id).first() is not None
     with Session(engine) as session:
         return session.query(model).filter_by(id=id).first() is not None
 
@@ -88,11 +90,12 @@ def load_full_wikipedia_dataset_sql():
     with Session(engine) as session:
         for page in tqdm(pages):
             id = page.find(id_tag).text
-            title = page.find(title_tag).text
-            text = page.find(text_tag).text
-            text_clean = get_clean_text(text)
-            standard_clean = get_standard_article(title)
-            if not _exists(id, model=WikiArticle):
+
+            if not _exists(id, model=WikiArticle, session=session):
+                title = page.find(title_tag).text
+                text = page.find(text_tag).text
+                text_clean = get_clean_text(text)
+                standard_clean = get_standard_article(title)
                 _save_article(id, title, standard_clean,
                               text_clean, session=session)
-        session.commit()
+                session.commit()

### 2023-03-28 main.py
move classifier logic and add evaluation

@@ -4,16 +4,19 @@ from dotenv import load_dotenv
 from src.make_dataset import make_dataset, write_dataset
 from src.experiment import analyze_dataset, plot_results
 
+from src.classifier.classifier import train_classifier
+
 load_dotenv()
 
 
 def main():
-    load_full_wikipedia_dataset_sql()
+    # load_full_wikipedia_dataset_sql()
     # load_simple_wiki_sql()
     # make_dataset()
     # write_dataset()
-    metrics, experiment = analyze_dataset()
-    plot_results(metrics, experiment)
+    # metrics, experiment = analyze_dataset()
+    # plot_results(metrics, experiment)
+    train_classifier()
 
 
 if __name__ == "__main__":

### 2023-03-28 src/classifier/classifier.py

@@ -1,0 +1,64 @@
+import torch
+from torch import nn
+from torch.optim import Adam
+from torch.utils.data import DataLoader
+import os
+
+from .model import Classifier
+from .params_setup import build_vocabulary
+from .load_data import get_train_test_ids, load_glove_data
+from .dataset import WikiArticleDataset
+from .device import device
+from .train import train
+from .evaluate import evaluate
+
+
+def train_classifier():
+    train_ids, test_ids = get_train_test_ids()
+    glove_word2idx, glove_idx2word, glove_embs = load_glove_data()
+    vocab_word2idx, vocab_idx2word, unk_idx, pad_idx, max_tokens = build_vocabulary(
+        train_ids, glove_word2idx, glove_idx2word)
+
+    train_dataset = WikiArticleDataset(
+        train_ids,
+        vocab_word2idx,
+        unk_idx, pad_idx,
+        pad_to_size=max_tokens,
+        sent_limit=5
+    )
+    test_dataset = WikiArticleDataset(
+        test_ids,
+        vocab_word2idx,
+        unk_idx, pad_idx,
+        pad_to_size=max_tokens,
+        sent_limit=1
+    )
+    train_dataloader = DataLoader(train_dataset, batch_size=64, shuffle=True)
+    test_dataloader = DataLoader(test_dataset, batch_size=10, shuffle=True)
+
+    model_path = "data/classifier_data/model.pt"
+
+    if os.path.exists(model_path):
+        model = Classifier(
+            glove_embs=glove_embs,
+            word2idx=vocab_word2idx,
+            idx2word=vocab_idx2word,
+            load_emb_weight=False
+        )
+        model.load_state_dict(torch.load(model_path))
+        model.to(device)
+
+    else:
+        model = Classifier(
+            glove_embs=glove_embs,
+            word2idx=vocab_word2idx,
+            idx2word=vocab_idx2word,
+            load_emb_weight=True
+        )
+        model.to(device)
+        optimizer = Adam(model.parameters(), lr=0.001)
+        criterion = nn.NLLLoss()
+        train(model, train_dataloader, optimizer, criterion, epochs=10)
+        torch.save(model.state_dict(), model_path)
+
+    evaluate(model, test_dataloader)

### 2023-03-28 src/classifier/dataset.py

@@ -1,0 +1,58 @@
+from typing import Dict, List, Tuple
+from torch.utils.data import Dataset
+
+from sqlalchemy.orm import Session
+from src.sql.models import WikiArticle
+from src.sql import engine
+from .preprocess import encode_text
+
+from .device import device
+import torch
+
+
+class WikiArticleDataset(Dataset):
+    def __init__(
+        self, ids: List[Tuple[int, str]],
+        word2idx: Dict[str, int],
+        unk_tok_idx: int,
+        pad_token_idx: int,
+        pad_to_size: int = 1000,
+        sent_limit: int = None
+    ) -> None:
+        super().__init__()
+        self._ids = ids
+        self.word2idx = word2idx
+        self.unk_idx = unk_tok_idx
+        self.pad_size = pad_to_size
+        self.pad_idx = pad_token_idx
+        self.sent_limit = sent_limit
+
+    def __len__(self):
+        return len(self._ids)
+
+    def __getitem__(self, index) -> Tuple[str, str]:
+        _id, level = self._ids[index]
+        with Session(engine) as session:
+            item = session.query(WikiArticle).filter_by(id=_id).first()
+            if level == "standard":
+                text = item.standard_text
+            else:
+                text = item.simple_text
+        tok_idx = encode_text(
+            text,
+            self.word2idx,
+            self.unk_idx,
+            limit_sents=self.sent_limit
+        )
+        tok_idx = torch.tensor(tok_idx, device=device, dtype=torch.long)
+
+        if len(tok_idx) < self.pad_size:
+            pad = torch.tensor(
+                [self.pad_idx] * (self.pad_size - len(tok_idx)), device=device, dtype=torch.long)
+            tok_idx = torch.cat([tok_idx, pad])
+        elif len(tok_idx) > self.pad_size:
+            tok_idx = tok_idx[:self.pad_size]
+
+        level = torch.tensor(
+            [1], device=device) if level == "standard" else torch.tensor([0], device=device)
+        return tok_idx, level

### 2023-03-28 src/classifier/device.py

@@ -1,0 +1,6 @@
+import torch
+
+if torch.backends.mps.is_available() and torch.backends.mps.is_built():
+    device = torch.device("cpu")
+else:
+    device = torch.device("cpu")

### 2023-03-28 src/classifier/evaluate.py

@@ -1,0 +1,16 @@
+from sklearn.metrics import precision_recall_fscore_support
+from torch.utils.data import DataLoader
+from .model import Classifier
+
+
+def evaluate(model: Classifier, test_dataloader: DataLoader):
+    y_true = []
+    y_pred = []
+    for document_batch, level_batch in test_dataloader:
+        out = model(document_batch)
+        y_true.extend(level_batch.reshape(-1).tolist())
+        out = out.argmax(dim=1).reshape(-1).tolist()
+        y_pred.extend(out)
+    precision, recall, fscore, support = precision_recall_fscore_support(
+        y_true, y_pred, average="macro")
+    print(f"Precision:\t{precision}\nRecall:\t{recall}\nF1\t{fscore}")

### 2023-03-28 src/classifier/load_data.py

@@ -1,0 +1,38 @@
+from typing import Dict, List
+from sqlalchemy.orm import Session
+from src.sql.models import WikiArticle
+from src.sql import engine
+import numpy as np
+
+
+def load_glove_data():
+    p = "data/glove.6B/glove.6B.50d.txt"
+    word2idx: Dict[str, int] = {}
+    idx2word: Dict[int, str] = {}
+    embs: List[List[float]] = []
+    with open(p, "r", encoding="utf-8") as file:
+        i = 0
+        for l in file:
+            l = l.split()
+            word, emb = l[0], l[1:]
+            emb = [float(x) for x in emb]
+            word2idx[word] = i
+            idx2word[i] = word
+            embs.append(emb)
+            i += 1
+
+    return word2idx, idx2word, np.array(embs)
+
+
+def get_train_test_ids(test_portion_size: float = 0.1):
+    with Session(engine) as session:
+        _ids = []
+        items = session.query(WikiArticle.id).limit(10_000).all()
+        for item in items:
+            _id = item[0]
+            _ids.append((_id, "standard"))
+            _ids.append((_id, "simple"))
+    train_size = 1 - test_portion_size
+    train = _ids[:int(len(_ids) * train_size)]
+    test = _ids[int(len(_ids) * train_size):]
+    return train, test

### 2023-03-28 src/classifier/model.py

@@ -1,0 +1,50 @@
+from torch import nn
+import torch
+from typing import Dict
+import numpy as np
+from .params_setup import get_embedding_weights
+
+
+class Classifier(nn.Module):
+    def __init__(
+        self,
+        hidden_size: int = 64,
+        glove_embs: np.ndarray = None,
+        word2idx: Dict[str, int] = None,
+        idx2word: Dict[int, str] = None,
+        load_emb_weight: bool = True
+    ) -> None:
+        super().__init__()
+        self.glove_embs = glove_embs
+        self.word2idx = word2idx
+        self.idx2word = idx2word
+
+        self.emb_size = 50
+        self.embedding = nn.Embedding(len(idx2word), self.emb_size)
+        if load_emb_weight:
+            self._load_embedding_weight()
+
+        self.lstm = nn.LSTM(
+            input_size=self.emb_size,
+            hidden_size=hidden_size,
+            num_layers=3,
+            batch_first=True,
+        )
+
+        self.dropout = nn.Dropout(0.1)
+        self.hidden_to_output = nn.Linear(hidden_size, 2)
+        self.softmax = nn.LogSoftmax(dim=1)
+
+    def _load_embedding_weight(self):
+        weight = get_embedding_weights(self.glove_embs, self.idx2word)
+        self.embedding.load_state_dict({"weight": weight})
+        self.embedding.weight.requires_grad = False
+
+    def forward(self, document_batch: torch.Tensor):
+        document_batch = self.embedding(document_batch)
+        lstm_out, _ = self.lstm(document_batch)
+        lstm_out = lstm_out[:, -1, :]
+        # lstm_out = self.dropout(lstm_out)
+        out = self.hidden_to_output(lstm_out)
+        out = self.softmax(out)
+        return out

### 2023-03-28 src/classifier/params_setup.py

@@ -1,0 +1,96 @@
+import pickle
+import os
+from typing import Dict, List, Tuple
+from tqdm import tqdm
+import numpy as np
+import torch
+
+from .util import random_vector
+from .device import device
+from sqlalchemy.orm import Session
+from src.sql.models import WikiArticle
+from src.sql import engine
+from .preprocess import clean_str
+from nltk.tokenize import word_tokenize
+
+
+def build_vocabulary(train_ids: List[Tuple[int, str]], glove_word2idx: Dict[str, int],  glove_idx2word: Dict[int, str]):
+    pickle_path = "data/classifier_data/vocabulary.pkl"
+
+    if os.path.exists(pickle_path):
+        with open(pickle_path, "rb") as file:
+            payload = pickle.load(file)
+            return payload["word2idx"], payload["idx2word"], payload["unk_idx"], payload["pad_idx"], payload["max_tokens"]
+
+    word2idx: Dict[str, int] = {}
+    idx2word = {}
+
+    for idx, word in glove_idx2word.items():
+        word2idx[word] = idx
+        idx2word[idx] = word
+
+    voc = set()
+    max_tokens = 0
+    with Session(engine) as session:
+        for _id, level in tqdm(train_ids):
+            doc = session.query(WikiArticle).filter_by(id=_id).first()
+            text = doc.simple_text if level == "simple" else doc.standard_text
+            text = clean_str(text)
+            tokens = word_tokenize(text)
+            for token in tokens:
+                voc.add(token)
+            max_tokens = max(max_tokens, len(tokens))
+
+    unprocessed = []
+    for word in voc:
+        glove_idx = glove_word2idx.get(word)
+        if glove_idx is None:
+            unprocessed.append(word)
+
+    i = len(word2idx)
+    for word in unprocessed:
+        while idx2word.get(i) is not None:
+            i += 1
+        word2idx[word] = i
+        idx2word[i] = word
+        i += 1
+
+    unk_idx = len(word2idx)
+    while idx2word.get(unk_idx) is not None:
+        unk_idx += 1
+    unk = "<UNK>"
+    word2idx[unk] = unk_idx
+    idx2word[unk_idx] = unk
+
+    pad_idx = len(word2idx)
+    while idx2word.get(pad_idx) is not None:
+        pad_idx += 1
+    pad = "<PAD>"
+    word2idx[pad] = pad_idx
+    idx2word[pad_idx] = pad
+
[…]

### 2023-03-28 src/classifier/preprocess.py

@@ -1,0 +1,27 @@
+import re
+from typing import Dict
+from nltk.tokenize import word_tokenize, sent_tokenize
+
+
+def clean_str(string):
+    # remove punctuation
+    string = re.sub(r"[^A-Za-z]", " ", string)
+    # remove extra spaces
+    string = re.sub(r"\s{2,}", " ", string)
+    return string.strip().lower()
+
+
+def encode_text(
+    string,
+    word2idx: Dict[str, int],
+    unk_idx: int,
+    limit_sents: int = None
+):
+    sents = sent_tokenize(string, language="english")
+    if limit_sents:
+        sents = sents[:limit_sents]
+    sents = ".".join(sents)
+    text = clean_str(sents)
+    toks = word_tokenize(text, language="english")
+    tok_idx = [word2idx.get(tok, unk_idx) for tok in toks]
+    return tok_idx

### 2023-03-28 src/classifier/train.py

@@ -1,0 +1,30 @@
+from .model import Classifier
+from torch import nn
+import torch
+from torch.utils.data import DataLoader
+from tqdm import tqdm
+
+
+def train(
+    model: Classifier,
+    train_loader: DataLoader,
+    optimizer: torch.optim.Optimizer,
+    criterion: nn.Module,
+    epochs: int = 10,
+):
+    for epoch in range(epochs):
+        print(f"Epoch {epoch + 1} of {epochs}")
+        pbar = tqdm(train_loader)
+        for (document_batch, level_batch) in pbar:
+            optimizer.zero_grad()
+            out = model(document_batch)
+            # combine losses
+            loss = 0
+            for i in range(out.shape[0]):
+                prediction = out[i].unsqueeze(0)
+                correct = level_batch[i]
+                loss += criterion(prediction, correct)
+            pbar.set_description(f"Loss: {loss.item()}")
+            loss.backward()
+
+            optimizer.step()

### 2023-03-28 src/classifier/util.py

@@ -1,0 +1,6 @@
+import torch
+from .device import device
+
+
+def random_vector(size: int) -> torch.Tensor:
+    return torch.rand(size, device=device, dtype=torch.float32)

### 2023-03-28 src/classifier/classifier.py
play around with model parameters

@@ -14,30 +14,33 @@ from .evaluate import evaluate
 
 
 def train_classifier():
+    print("Getting ids")
     train_ids, test_ids = get_train_test_ids()
+    print("Loading glove data")
     glove_word2idx, glove_idx2word, glove_embs = load_glove_data()
+    print("Building vocabulary")
     vocab_word2idx, vocab_idx2word, unk_idx, pad_idx, max_tokens = build_vocabulary(
         train_ids, glove_word2idx, glove_idx2word)
 
+    print("Loading datasets")
     train_dataset = WikiArticleDataset(
         train_ids,
         vocab_word2idx,
         unk_idx, pad_idx,
-        pad_to_size=max_tokens,
-        sent_limit=5
+        pad_to_size=2000,
     )
     test_dataset = WikiArticleDataset(
         test_ids,
         vocab_word2idx,
         unk_idx, pad_idx,
-        pad_to_size=max_tokens,
-        sent_limit=1
+        pad_to_size=10000,
     )
     train_dataloader = DataLoader(train_dataset, batch_size=64, shuffle=True)
     test_dataloader = DataLoader(test_dataset, batch_size=10, shuffle=True)
 
     model_path = "data/classifier_data/model.pt"
 
+    print("Loading model")
     if os.path.exists(model_path):
         model = Classifier(
             glove_embs=glove_embs,
@@ -45,7 +48,9 @@ def train_classifier():
             idx2word=vocab_idx2word,
             load_emb_weight=False
         )
+        print("Loading model weights")
         model.load_state_dict(torch.load(model_path))
+        print("Model loaded, moving to device")
         model.to(device)
 
     else:
@@ -58,7 +63,8 @@ def train_classifier():
         model.to(device)
         optimizer = Adam(model.parameters(), lr=0.001)
         criterion = nn.NLLLoss()
-        train(model, train_dataloader, optimizer, criterion, epochs=10)
+        train(model, train_dataloader, optimizer, criterion, epochs=5)
         torch.save(model.state_dict(), model_path)
 
+    print("Evaluating model")
     evaluate(model, test_dataloader)

### 2023-03-28 src/classifier/evaluate.py

@@ -1,12 +1,13 @@
 from sklearn.metrics import precision_recall_fscore_support
 from torch.utils.data import DataLoader
 from .model import Classifier
+from tqdm import tqdm
 
 
 def evaluate(model: Classifier, test_dataloader: DataLoader):
     y_true = []
     y_pred = []
-    for document_batch, level_batch in test_dataloader:
+    for document_batch, level_batch in tqdm(test_dataloader):
         out = model(document_batch)
         y_true.extend(level_batch.reshape(-1).tolist())
         out = out.argmax(dim=1).reshape(-1).tolist()

### 2023-03-28 src/classifier/load_data.py

@@ -27,7 +27,7 @@ def load_glove_data():
 def get_train_test_ids(test_portion_size: float = 0.1):
     with Session(engine) as session:
         _ids = []
-        items = session.query(WikiArticle.id).limit(10_000).all()
+        items = session.query(WikiArticle.id).limit(999999).all()
         for item in items:
             _id = item[0]
             _ids.append((_id, "standard"))

### 2023-03-28 src/classifier/params_setup.py

@@ -41,19 +41,19 @@ def build_vocabulary(train_ids: List[Tuple[int, str]], glove_word2idx: Dict[str,
                 voc.add(token)
             max_tokens = max(max_tokens, len(tokens))
 
-    unprocessed = []
-    for word in voc:
-        glove_idx = glove_word2idx.get(word)
-        if glove_idx is None:
-            unprocessed.append(word)
+    # unprocessed = []
+    # for word in voc:
+    #     glove_idx = glove_word2idx.get(word)
+    #     if glove_idx is None:
+    #         unprocessed.append(word)
 
-    i = len(word2idx)
-    for word in unprocessed:
-        while idx2word.get(i) is not None:
-            i += 1
-        word2idx[word] = i
-        idx2word[i] = word
-        i += 1
+    # i = len(word2idx)
+    # for word in unprocessed:
+    #     while idx2word.get(i) is not None:
+    #         i += 1
+    #     word2idx[word] = i
+    #     idx2word[i] = word
+    #     i += 1
 
     unk_idx = len(word2idx)
     while idx2word.get(unk_idx) is not None:

### 2023-03-28 main.py
make better classifier dataset

@@ -5,6 +5,7 @@ from src.make_dataset import make_dataset, write_dataset
 from src.experiment import analyze_dataset, plot_results
 
 from src.classifier.classifier import train_classifier
+from src.classifier.make_dataset import make_classifier_dataset
 
 load_dotenv()
 
@@ -16,7 +17,8 @@ def main():
     # write_dataset()
     # metrics, experiment = analyze_dataset()
     # plot_results(metrics, experiment)
-    train_classifier()
+    # train_classifier()
+    make_classifier_dataset()
 
 
 if __name__ == "__main__":

### 2023-03-28 src/classifier/classifier.py

@@ -33,7 +33,7 @@ def train_classifier():
         test_ids,
         vocab_word2idx,
         unk_idx, pad_idx,
-        pad_to_size=10000,
+        pad_to_size=2000,
     )
     train_dataloader = DataLoader(train_dataset, batch_size=64, shuffle=True)
     test_dataloader = DataLoader(test_dataset, batch_size=10, shuffle=True)

### 2023-03-28 src/classifier/make_dataset.py

@@ -1,0 +1,65 @@
+from src.sql.models import WikiArticle, ClassifierDataset
+from src.sql import engine
+import random
+from tqdm import tqdm
+
+from sqlalchemy.orm.session import Session
+
+
+def _get_paragraphs(text: str):
+    paragraphs = text.split("\n\n")
+    paragraphs = [paragraph.strip()
+                  for paragraph in paragraphs if len(paragraph) > 0 and "*" not in paragraph]
+    return paragraphs
+
+
+def make_classifier_dataset():
+    with Session(engine) as session:
+        #     len_objects = session.query(ClassifierDataset).count()
+        #     q = session.query(WikiArticle)
+        #     for article in tqdm(q, total=len_objects):
+        #         paragraphs_simple = _get_paragraphs(article.simple_text)
+        #         paragraphs_standard = _get_paragraphs(article.standard_text)
+        #         for p in paragraphs_simple:
+        #             session.add(ClassifierDataset(
+        #                 partition="train",
+        #                 level="simple",
+        #                 text=p,
+        #             ))
+        #         for p in paragraphs_standard:
+        #             session.add(ClassifierDataset(
+        #                 partition="train",
+        #                 level="standard",
+        #                 text=p,
+        #             ))
+        #     session.commit()
+
+        _ids = session.query(ClassifierDataset.id).all()
+
+        for _id in _ids:
+            _id = _id[0]
+            session.query(ClassifierDataset).filter(
+                ClassifierDataset.id == _id).update({"partition": "train"})
+
+        session.commit()
+
+        len_test = 20_000
+
+        simple_ids = session.query(ClassifierDataset.id).filter(
+            ClassifierDataset.level == "simple").all()
+        standard_ids = session.query(ClassifierDataset.id).filter(
+            ClassifierDataset.level == "standard").all()
+
+        simple_ids = [id[0] for id in simple_ids]
+        standard_ids = [id[0] for id in standard_ids]
+        random.shuffle(simple_ids)
+        random.shuffle(standard_ids)
+
+        test_ids_simple = simple_ids[:len_test]
+        test_ids_standard = standard_ids[:len_test]
+
+        for _id in [*test_ids_simple, *test_ids_standard]:
+            session.query(ClassifierDataset).filter(
+                ClassifierDataset.id == _id).update({"partition": "test"})
+
+        session.commit()

### 2023-03-28 src/sql/models.py

@@ -46,6 +46,15 @@ class ClearCorpusEntry(Base):
     cml2ri: Mapped[Optional[float]]
 
 
+class ClassifierDataset(Base):
+    __tablename__ = "classifier_dataset"
+
+    id: Mapped[int] = mapped_column(primary_key=True)
+    partition: Mapped[str]
+    text: Mapped[str]
+    level: Mapped[str]
+
+
 class WikiArticle(Base):
     __tablename__ = "wiki_article"
 

### 2023-03-28 src/classifier/classifier.py
fine-grain classifier dataset

@@ -21,19 +21,19 @@ def train_classifier():
     print("Building vocabulary")
     vocab_word2idx, vocab_idx2word, unk_idx, pad_idx, max_tokens = build_vocabulary(
         train_ids, glove_word2idx, glove_idx2word)
-
+    print("max_tokens", max_tokens)
     print("Loading datasets")
     train_dataset = WikiArticleDataset(
         train_ids,
         vocab_word2idx,
         unk_idx, pad_idx,
-        pad_to_size=2000,
+        pad_to_size=max_tokens,
     )
     test_dataset = WikiArticleDataset(
         test_ids,
         vocab_word2idx,
         unk_idx, pad_idx,
-        pad_to_size=2000,
+        pad_to_size=max_tokens,
     )
     train_dataloader = DataLoader(train_dataset, batch_size=64, shuffle=True)
     test_dataloader = DataLoader(test_dataset, batch_size=10, shuffle=True)

### 2023-03-28 src/classifier/dataset.py

@@ -2,7 +2,7 @@ from typing import Dict, List, Tuple
 from torch.utils.data import Dataset
 
 from sqlalchemy.orm import Session
-from src.sql.models import WikiArticle
+from src.sql.models import ClassifierDataset
 from src.sql import engine
 from .preprocess import encode_text
 
@@ -33,11 +33,9 @@ class WikiArticleDataset(Dataset):
     def __getitem__(self, index) -> Tuple[str, str]:
         _id, level = self._ids[index]
         with Session(engine) as session:
-            item = session.query(WikiArticle).filter_by(id=_id).first()
-            if level == "standard":
-                text = item.standard_text
-            else:
-                text = item.simple_text
+            item = session.query(ClassifierDataset).filter_by(id=_id).first()
+            text = item.text
+            print(text)
         tok_idx = encode_text(
             text,
             self.word2idx,

### 2023-03-28 src/classifier/load_data.py

@@ -1,6 +1,6 @@
 from typing import Dict, List
 from sqlalchemy.orm import Session
-from src.sql.models import WikiArticle
+from src.sql.models import ClassifierDataset
 from src.sql import engine
 import numpy as np
 
@@ -24,15 +24,24 @@ def load_glove_data():
     return word2idx, idx2word, np.array(embs)
 
 
-def get_train_test_ids(test_portion_size: float = 0.1):
+def get_train_test_ids():
+    train, test = [], []
+    train_limit = 200_000
     with Session(engine) as session:
-        _ids = []
-        items = session.query(WikiArticle.id).limit(999999).all()
-        for item in items:
-            _id = item[0]
-            _ids.append((_id, "standard"))
-            _ids.append((_id, "simple"))
-    train_size = 1 - test_portion_size
-    train = _ids[:int(len(_ids) * train_size)]
-    test = _ids[int(len(_ids) * train_size):]
+        q_simple_train = session.query(ClassifierDataset).filter_by(
+            level="simple", partition="train").limit(train_limit)
+        q_standard_train = session.query(ClassifierDataset).filter_by(
+            level="standard", partition="train").limit(train_limit)
+        q_simple_test = session.query(ClassifierDataset).filter_by(
+            level="simple", partition="test")
+        q_standard_test = session.query(ClassifierDataset).filter_by(
+            level="standard", partition="test")
+        for item in q_simple_train:
+            train.append((item.id, "simple"))
+        for item in q_standard_train:
+            train.append((item.id, "standard"))
+        for item in q_simple_test:
+            test.append((item.id, "simple"))
+        for item in q_standard_test:
+            test.append((item.id, "standard"))
     return train, test

### 2023-03-28 src/classifier/make_dataset.py

@@ -8,40 +8,50 @@ from sqlalchemy.orm.session import Session
 
 def _get_paragraphs(text: str):
     paragraphs = text.split("\n\n")
-    paragraphs = [paragraph.strip()
-                  for paragraph in paragraphs if len(paragraph) > 0 and "*" not in paragraph]
+    paragraphs = [
+        paragraph.strip() for paragraph in paragraphs
+        if len(paragraph.split()) > 20
+        and len(paragraph.split()) < 2000
+        and "*" not in paragraph
+        and "<" not in paragraph
+        and ">" not in paragraph
+        and "==" not in paragraph
+        and "{" not in paragraph
+        and "colspan=" not in paragraph
+        and "style=" not in paragraph
+    ]
     return paragraphs
 
 
 def make_classifier_dataset():
     with Session(engine) as session:
-        #     len_objects = session.query(ClassifierDataset).count()
-        #     q = session.query(WikiArticle)
-        #     for article in tqdm(q, total=len_objects):
-        #         paragraphs_simple = _get_paragraphs(article.simple_text)
-        #         paragraphs_standard = _get_paragraphs(article.standard_text)
-        #         for p in paragraphs_simple:
-        #             session.add(ClassifierDataset(
-        #                 partition="train",
-        #                 level="simple",
-        #                 text=p,
-        #             ))
-        #         for p in paragraphs_standard:
-        #             session.add(ClassifierDataset(
-        #                 partition="train",
-        #                 level="standard",
-        #                 text=p,
-        #             ))
-        #     session.commit()
-
-        _ids = session.query(ClassifierDataset.id).all()
-
-        for _id in _ids:
-            _id = _id[0]
-            session.query(ClassifierDataset).filter(
-                ClassifierDataset.id == _id).update({"partition": "train"})
-
+        len_objects = session.query(ClassifierDataset).count()
+        q = session.query(WikiArticle)
+        for article in tqdm(q, total=len_objects):
+            paragraphs_simple = _get_paragraphs(article.simple_text)
+            paragraphs_standard = _get_paragraphs(article.standard_text)
+            for p in paragraphs_simple:
+                session.add(ClassifierDataset(
+                    partition="train",
+                    level="simple",
+                    text=p,
+                ))
+            for p in paragraphs_standard:
+                session.add(ClassifierDataset(
+                    partition="train",
+                    level="standard",
+                    text=p,
+                ))
         session.commit()
+
[…]

### 2023-03-28 src/classifier/params_setup.py

@@ -8,7 +8,7 @@ import torch
 from .util import random_vector
 from .device import device
 from sqlalchemy.orm import Session
-from src.sql.models import WikiArticle
+from src.sql.models import ClassifierDataset
 from src.sql import engine
 from .preprocess import clean_str
 from nltk.tokenize import word_tokenize
@@ -33,8 +33,8 @@ def build_vocabulary(train_ids: List[Tuple[int, str]], glove_word2idx: Dict[str,
     max_tokens = 0
     with Session(engine) as session:
         for _id, level in tqdm(train_ids):
-            doc = session.query(WikiArticle).filter_by(id=_id).first()
-            text = doc.simple_text if level == "simple" else doc.standard_text
+            doc = session.query(ClassifierDataset).filter_by(id=_id).first()
+            text = doc.text
             text = clean_str(text)
             tokens = word_tokenize(text)
             for token in tokens:

### 2023-03-28 main.py
update classifier defs

@@ -17,8 +17,8 @@ def main():
     # write_dataset()
     # metrics, experiment = analyze_dataset()
     # plot_results(metrics, experiment)
-    # train_classifier()
-    make_classifier_dataset()
+    train_classifier()
+    # make_classifier_dataset()
 
 
 if __name__ == "__main__":

### 2023-03-28 src/classifier/classifier.py

@@ -15,13 +15,13 @@ from .evaluate import evaluate
 
 def train_classifier():
     print("Getting ids")
-    train_ids, test_ids = get_train_test_ids()
+    train_ids, test_ids = get_train_test_ids(train_limit=10_000)
     print("Loading glove data")
     glove_word2idx, glove_idx2word, glove_embs = load_glove_data()
     print("Building vocabulary")
     vocab_word2idx, vocab_idx2word, unk_idx, pad_idx, max_tokens = build_vocabulary(
         train_ids, glove_word2idx, glove_idx2word)
-    print("max_tokens", max_tokens)
+    print(f"Longest entry has {max_tokens} tokens")
     print("Loading datasets")
     train_dataset = WikiArticleDataset(
         train_ids,
@@ -36,7 +36,7 @@ def train_classifier():
         pad_to_size=max_tokens,
     )
     train_dataloader = DataLoader(train_dataset, batch_size=64, shuffle=True)
-    test_dataloader = DataLoader(test_dataset, batch_size=10, shuffle=True)
+    test_dataloader = DataLoader(test_dataset, batch_size=64, shuffle=True)
 
     model_path = "data/classifier_data/model.pt"
 

### 2023-03-28 src/classifier/dataset.py

@@ -35,7 +35,6 @@ class WikiArticleDataset(Dataset):
         with Session(engine) as session:
             item = session.query(ClassifierDataset).filter_by(id=_id).first()
             text = item.text
-            print(text)
         tok_idx = encode_text(
             text,
             self.word2idx,

### 2023-03-28 src/classifier/load_data.py

@@ -24,9 +24,8 @@ def load_glove_data():
     return word2idx, idx2word, np.array(embs)
 
 
-def get_train_test_ids():
+def get_train_test_ids(train_limit: int = 200_000):
     train, test = [], []
-    train_limit = 200_000
     with Session(engine) as session:
         q_simple_train = session.query(ClassifierDataset).filter_by(
             level="simple", partition="train").limit(train_limit)

### 2023-03-28 paper/paper.tex
move classifier logic and add evaluation
play around

@@
 
+\input{figs/lldl_means.tex}
+
+LLDL (length of longest dependency link) is another metric that performs well in terms of level differentiation.
+It also reflects the common belief that the biggest difficulty spike in language learning - as shown by Figure~\ref{fig:lldl-mean}.
+As with most metrics, however, the natural progression of the CEFR levels is lost, particularly in the middle portions.
+
 \section{Discussion and future research}
+
+This work attempted to gauge true readability levels of GPT-3 output texts with prompts across 10 topics and CEFR proficiency levels, optimising thus for the second language learning application environment.
+Outputs computed from the metrics set suggest that GPT-3 is at least somewhat aware of syntax and lexis when performin generation for a target readability level.
+
+Future research into this topic could explore a wider variety of textual metrics, as this work only selected a handful that lent themselves particularly well to interpretation and did not require steep hardware requirements to obtain.
 

### 2023-03-28 src/classifier/classifier.py

@@ -15,7 +15,7 @@ from .evaluate import evaluate
 
 def train_classifier():
     print("Getting ids")
-    train_ids, test_ids = get_train_test_ids(train_limit=10_000)
+    train_ids, test_ids = get_train_test_ids(train_limit=1000)
     print("Loading glove data")
     glove_word2idx, glove_idx2word, glove_embs = load_glove_data()
     print("Building vocabulary")
@@ -61,7 +61,7 @@ def train_classifier():
             load_emb_weight=True
         )
         model.to(device)
-        optimizer = Adam(model.parameters(), lr=0.001)
+        optimizer = Adam(model.parameters(), lr=1e-6)
         criterion = nn.NLLLoss()
         train(model, train_dataloader, optimizer, criterion, epochs=5)
         torch.save(model.state_dict(), model_path)

### 2023-03-28 src/classifier/device.py

@@ -1,6 +1,6 @@
 import torch
 
 if torch.backends.mps.is_available() and torch.backends.mps.is_built():
-    device = torch.device("mps")
+    device = torch.device("cpu")
 else:
     device = torch.device("cpu")

### 2023-03-28 src/classifier/evaluate.py

@@ -2,16 +2,18 @@ from sklearn.metrics import precision_recall_fscore_support
 from torch.utils.data import DataLoader
 from .model import Classifier
 from tqdm import tqdm
+import torch
 
 
 def evaluate(model: Classifier, test_dataloader: DataLoader):
     y_true = []
     y_pred = []
-    for document_batch, level_batch in tqdm(test_dataloader):
-        out = model(document_batch)
-        y_true.extend(level_batch.reshape(-1).tolist())
-        out = out.argmax(dim=1).reshape(-1).tolist()
-        y_pred.extend(out)
-    precision, recall, fscore, support = precision_recall_fscore_support(
-        y_true, y_pred, average="macro")
-    print(f"Precision:\t{precision}\nRecall:\t{recall}\nF1\t{fscore}")
+    with torch.no_grad():
+        for document_batch, level_batch in tqdm(test_dataloader):
+            out = model(document_batch)
+            y_true.extend(level_batch.reshape(-1).tolist())
+            out = out.argmax(dim=1).reshape(-1).tolist()
+            y_pred.extend(out)
+        precision, recall, fscore, support = precision_recall_fscore_support(
+            y_true, y_pred, average="macro")
+        print(f"Precision:\t{precision}\nRecall:\t{recall}\nF1\t{fscore}")

### 2023-03-28 src/classifier/model.py

@@ -8,7 +8,7 @@ from .params_setup import get_embedding_weights
 class Classifier(nn.Module):
     def __init__(
         self,
-        hidden_size: int = 64,
+        hidden_size: int = 256,
         glove_embs: np.ndarray = None,
         word2idx: Dict[str, int] = None,
         idx2word: Dict[int, str] = None,
@@ -38,13 +38,19 @@ class Classifier(nn.Module):
     def _load_embedding_weight(self):
         weight = get_embedding_weights(self.glove_embs, self.idx2word)
         self.embedding.load_state_dict({"weight": weight})
-        self.embedding.weight.requires_grad = False
+        self.embedding.weight.requires_grad = True
 
     def forward(self, document_batch: torch.Tensor):
         document_batch = self.embedding(document_batch)
+        # print("document_batch")
+        # print(document_batch[0][0])
+        # print(document_batch[1][0])
         lstm_out, _ = self.lstm(document_batch)
         lstm_out = lstm_out[:, -1, :]
-        # lstm_out = self.dropout(lstm_out)
+        # print("lstm_out")
+        # print(lstm_out[0])
+        # print(lstm_out[1])
+        lstm_out = self.dropout(lstm_out)
         out = self.hidden_to_output(lstm_out)
         out = self.softmax(out)
         return out

### 2023-03-28 src/classifier/params_setup.py

@@ -41,19 +41,19 @@ def build_vocabulary(train_ids: List[Tuple[int, str]], glove_word2idx: Dict[str,
                 voc.add(token)
             max_tokens = max(max_tokens, len(tokens))
 
-    # unprocessed = []
-    # for word in voc:
-    #     glove_idx = glove_word2idx.get(word)
-    #     if glove_idx is None:
-    #         unprocessed.append(word)
+    unprocessed = []
+    for word in voc:
+        glove_idx = glove_word2idx.get(word)
+        if glove_idx is None:
+            unprocessed.append(word)
 
-    # i = len(word2idx)
-    # for word in unprocessed:
-    #     while idx2word.get(i) is not None:
-    #         i += 1
-    #     word2idx[word] = i
-    #     idx2word[i] = word
-    #     i += 1
+    i = len(word2idx)
+    for word in unprocessed:
+        while idx2word.get(i) is not None:
+            i += 1
+        word2idx[word] = i
+        idx2word[i] = word
+        i += 1
 
     unk_idx = len(word2idx)
     while idx2word.get(unk_idx) is not None:

### 2023-03-28 src/classifier/classifier.py
play around

@@ -15,7 +15,7 @@ from .evaluate import evaluate
 
 def train_classifier():
     print("Getting ids")
-    train_ids, test_ids = get_train_test_ids(train_limit=1000)
+    train_ids, test_ids = get_train_test_ids(train_limit=140_000)
     print("Loading glove data")
     glove_word2idx, glove_idx2word, glove_embs = load_glove_data()
     print("Building vocabulary")
@@ -61,7 +61,7 @@ def train_classifier():
             load_emb_weight=True
         )
         model.to(device)
-        optimizer = Adam(model.parameters(), lr=1e-6)
+        optimizer = Adam(model.parameters(), lr=0.005)
         criterion = nn.NLLLoss()
         train(model, train_dataloader, optimizer, criterion, epochs=5)
         torch.save(model.state_dict(), model_path)

### 2023-03-28 src/classifier/device.py

@@ -1,6 +1,6 @@
 import torch
 
 if torch.backends.mps.is_available() and torch.backends.mps.is_built():
-    device = torch.device("cpu")
+    device = torch.device("mps")
 else:
     device = torch.device("cpu")

### 2023-03-28 src/classifier/model.py

@@ -38,7 +38,7 @@ class Classifier(nn.Module):
     def _load_embedding_weight(self):
         weight = get_embedding_weights(self.glove_embs, self.idx2word)
         self.embedding.load_state_dict({"weight": weight})
-        self.embedding.weight.requires_grad = True
+        self.embedding.weight.requires_grad = False
 
     def forward(self, document_batch: torch.Tensor):
         document_batch = self.embedding(document_batch)

### 2023-03-28 src/classifier/train.py

@@ -3,6 +3,7 @@ from torch import nn
 import torch
 from torch.utils.data import DataLoader
 from tqdm import tqdm
+import numpy as np
 
 
 def train(
@@ -12,6 +13,7 @@ def train(
     criterion: nn.Module,
     epochs: int = 10,
 ):
+    model.train()
     for epoch in range(epochs):
         print(f"Epoch {epoch + 1} of {epochs}")
         pbar = tqdm(train_loader)
@@ -19,11 +21,8 @@ def train(
             optimizer.zero_grad()
             out = model(document_batch)
             # combine losses
-            loss = 0
-            for i in range(out.shape[0]):
-                prediction = out[i].unsqueeze(0)
-                correct = level_batch[i]
-                loss += criterion(prediction, correct)
+            loss = criterion(out, level_batch.reshape(-1))
+
             pbar.set_description(f"Loss: {loss.item()}")
             loss.backward()
 

### 2023-03-28 src/classifier/classifier.py
play around with classifier

@@ -15,7 +15,7 @@ from .evaluate import evaluate
 
 def train_classifier():
     print("Getting ids")
-    train_ids, test_ids = get_train_test_ids(train_limit=140_000)
+    train_ids, test_ids = get_train_test_ids(train_limit=1000)
     print("Loading glove data")
     glove_word2idx, glove_idx2word, glove_embs = load_glove_data()
     print("Building vocabulary")
@@ -63,7 +63,8 @@ def train_classifier():
         model.to(device)
         optimizer = Adam(model.parameters(), lr=0.005)
         criterion = nn.NLLLoss()
-        train(model, train_dataloader, optimizer, criterion, epochs=5)
+        train(model, train_dataloader, optimizer,
+              criterion, epochs=5, pad_idx=pad_idx)
         torch.save(model.state_dict(), model_path)
 
     print("Evaluating model")

### 2023-03-28 src/classifier/model.py

@@ -36,9 +36,10 @@ class Classifier(nn.Module):
         self.softmax = nn.LogSoftmax(dim=1)
 
     def _load_embedding_weight(self):
-        weight = get_embedding_weights(self.glove_embs, self.idx2word)
+        weight = get_embedding_weights(
+            self.glove_embs, self.idx2word, self.word2idx)
         self.embedding.load_state_dict({"weight": weight})
-        self.embedding.weight.requires_grad = False
+        self.embedding.weight.requires_grad = True
 
     def forward(self, document_batch: torch.Tensor):
         document_batch = self.embedding(document_batch)

### 2023-03-28 src/classifier/params_setup.py

@@ -13,6 +13,9 @@ from src.sql import engine
 from .preprocess import clean_str
 from nltk.tokenize import word_tokenize
 
+PAD_TOKEN = "<PAD>"
+UNK_TOKEN = "<UNK>"
+
 
 def build_vocabulary(train_ids: List[Tuple[int, str]], glove_word2idx: Dict[str, int],  glove_idx2word: Dict[int, str]):
     pickle_path = "data/classifier_data/vocabulary.pkl"
@@ -58,14 +61,14 @@ def build_vocabulary(train_ids: List[Tuple[int, str]], glove_word2idx: Dict[str,
     unk_idx = len(word2idx)
     while idx2word.get(unk_idx) is not None:
         unk_idx += 1
-    unk = "<UNK>"
+    unk = UNK_TOKEN
     word2idx[unk] = unk_idx
     idx2word[unk_idx] = unk
 
     pad_idx = len(word2idx)
     while idx2word.get(pad_idx) is not None:
         pad_idx += 1
-    pad = "<PAD>"
+    pad = PAD_TOKEN
     word2idx[pad] = pad_idx
     idx2word[pad_idx] = pad
 
@@ -83,7 +86,8 @@ def build_vocabulary(train_ids: List[Tuple[int, str]], glove_word2idx: Dict[str,
     return word2idx, idx2word, unk_idx, pad_idx, max_tokens
 
 
-def get_embedding_weights(embs: np.ndarray, idx2word: Dict[int, str]):
+def get_embedding_weights(embs: np.ndarray, idx2word: Dict[int, str], word2idx: Dict[str, int]):
+    pad_idx = word2idx[PAD_TOKEN]
     embedding_dim = 50
     vocab_size = len(idx2word)
     weight_matrix = torch.zeros((vocab_size, embedding_dim))
@@ -91,6 +95,9 @@ def get_embedding_weights(embs: np.ndarray, idx2word: Dict[int, str]):
         if idx < len(embs):
             weight_matrix[idx] = torch.tensor(
                 embs[idx], device=device, dtype=torch.float32)
+        elif idx == pad_idx:
+            weight_matrix[idx] = torch.zeros(
+                embedding_dim, device=device, dtype=torch.float32)
         else:
             weight_matrix[idx] = random_vector(embedding_dim)
     return weight_matrix

### 2023-03-28 src/classifier/train.py

@@ -4,6 +4,7 @@ import torch
 from torch.utils.data import DataLoader
 from tqdm import tqdm
 import numpy as np
+from .util import de_pad
 
 
 def train(
@@ -12,18 +13,31 @@ def train(
     optimizer: torch.optim.Optimizer,
     criterion: nn.Module,
     epochs: int = 10,
+    pad_idx: int = 0
 ):
     model.train()
     for epoch in range(epochs):
         print(f"Epoch {epoch + 1} of {epochs}")
         pbar = tqdm(train_loader)
         for (document_batch, level_batch) in pbar:
-            optimizer.zero_grad()
-            out = model(document_batch)
-            # combine losses
-            loss = criterion(out, level_batch.reshape(-1))
+            total_losses = 0
+            for document, level in zip(document_batch, level_batch):
+                optimizer.zero_grad()
+                document = de_pad(document, pad_idx).unsqueeze(0)
+                out = model(document)
+                loss = criterion(out, level.reshape(-1))
+                loss.backward()
+                optimizer.step()
+                total_losses += loss.item()
+            total_losses = total_losses / len(document_batch)
+            pbar.set_description(f"Loss: {total_losses}")
 
-            pbar.set_description(f"Loss: {loss.item()}")
-            loss.backward()
+            # optimizer.zero_grad()
+            # out = model(document_batch)
+            # # combine losses
+            # loss = criterion(out, level_batch.reshape(-1))
 
-            optimizer.step()
+            # pbar.set_description(f"Loss: {loss.item()}")
+            # loss.backward()
+
+            # optimizer.step()

### 2023-03-28 src/classifier/util.py

@@ -4,3 +4,13 @@ from .device import device
 
 def random_vector(size: int) -> torch.Tensor:
     return torch.rand(size, device=device, dtype=torch.float32)
+
+
+def de_pad(tensor: torch.Tensor, pad_idx: int):
+    t_new = []
+    for t in tensor:
+        if t.item() == pad_idx:
+            break
+        else:
+            t_new.append(t.item())
+    return torch.tensor(t_new, device=device, dtype=torch.long)

### 2023-03-28 paper/paper.tex
play around
update paper

@@
     This paper examines readability-related metrics of AI-generated text, specifically texts produced by OpenAI's GPT-3.
-Generation was performed across 10 topics and the 6 proficiency levels defined by the CEFR framework, and outputs were compared with Simple Wikipedia and (standard) Wikipedia articles on the same topics.
+Generation was performed for the English language across 10 topics and the 6 proficiency levels defined by the CEFR framework, and outputs were compared with Simple Wikipedia and (standard) Wikipedia articles on the same topics.
+7 commonly used complexity metrics have been employed, targeting chiefly lexis and syntax.
+In the limited scope of this experiment, GPT-3 is observed to at least partially aware of what constitutes reading difficulty.
 \end{abstract}
@@
 Owing to the increased popularity of these tools, more research is desirable on their applicability in a number of different settings.
+For example, GPT-3 has been observed to "make up" information about some real-world events or phenomena, a fact that makes it hard to rely on for teaching.
+In the specific context of this experiment, employing GPT-3 for language-learning material is only valuable insofar as the generated content is adequate in difficulty to the audience.
 
-This work aims to make a contribution to those questions by verifying the readability level of text content generated by GPT-3 with regard to 10 topics and the 6 proficiency levels defined by the CEFR framework.
+This work aims to make a contribution to these questions by verifying the readability level of text content generated by GPT-3 with regard to 10 topics and the 6 proficiency levels defined by the CEFR framework.
 A number of complexity metrics are computed on the generated texts and compared with Simple Wikipedia\footnote{\url{https://simple.wikipedia.org/wiki/Main_Page}} as well as Wikipedia\footnote{\url{https://www.wikipedia.org/}} articles on the same topics in order to gauge the true readability levels of the model outputs.
@@
 
+Interesting discussions can be had around topics as well, since, depending on GPT-3's ability to generalize across domains, certain areas might be problem areas for it.
+For example, "semantics" and "quantum mechanics" would be hard to write about for A1 level students as they are likely to be children who have not interacted with those fields yet.
+
+Generally, a difference in values is visible across the board, more so at the lower difficuly levels.
+At CEFR levels B2 and above, many of the metrics appear similar regardless of topic.
+Among the selected measures, NVR (noun to verb ratio) shows the highest cross-topic variation, maintaining strong differentitation even at the C2-level texts.
+Figure \ref{fig:wit-nvr-topics} shows metric values at the A1 and C2 levels, which indicate that topics commonly regarded as intrinsically have higher information density in GPT-3 texts.
+This is however a weak case, as few metrics differ so sharply as NVR, and bars in these figures correspond to a single text entry.
+
+\input{figs/nvr_topics.tex}
+
 \section{Discussion and future research}
@@
 Future research into this topic could explore a wider variety of textual metrics, as this work only selected a handful that lent themselves particularly well to interpretation and did not require steep hardware requirements to obtain.
+
+Since recent research (e.g.
+\citealp{Deutsch_2020}) has also focused on readability assessment through neural approaches without such metrics, this could also be explored with AI text data.
+An attempt was made in relationship to this experiment to train a readability analyzer with GLoVe word embeddings \citep{pennington2014glove} and Wikipedia data, but this unfortunaly was inconclusive, and thus results from this analyzer haven't been reported.
+
+Mainly, however, future extensions of this experiment or related experiments would benefit from higher volume.
+Simply put, more data is necessary across multiple domains to make educated conclusions about the quality of GPT-3 outputs in a teaching setting.
+While data is easy to generate, quality annotations on it are yet unavailable.
+Indeed, it is a good question whether making any is a fruitful enterprise, as the actual models behind GPT-3 and ChatGPT are constantly updated - thus making it hard to pinpoint what exactly is being studied.
 

### 2023-03-28 main.py

@@ -15,9 +15,9 @@ def main():
     # load_simple_wiki_sql()
     # make_dataset()
     # write_dataset()
-    # metrics, experiment = analyze_dataset()
-    # plot_results(metrics, experiment)
-    train_classifier()
+    metrics, experiment = analyze_dataset()
+    plot_results(metrics, experiment)
+    # train_classifier()
     # make_classifier_dataset()
 
 

### 2023-03-28 src/experiment/plot_results.py

@@ -52,6 +52,29 @@ def _make_topic_metric_plot(metric: str, topic: str, experiment: Experiment, sho
     plt.cla()
 
 
+def _plot_levels(metrics: List[Metric], experiment: Experiment):
+    def _name(string):
+        if len(string) > 10:
+            return string[:10] + "..."
+        return string
+
+    for metric in metrics:
+        for level in LEVELS:
+            values = experiment.results.filter(
+                metrics=[metric.name], levels=[level]).all()
+            tups = [(value["topic"], value["value"]) for value in values]
+            tups.sort(key=lambda tup: tup[1])
+            x = list(range(len(tups)))
+            height = [tup[1] for tup in tups]
+            labels = [_name(tup[0]) for tup in tups]
+            plt.bar(x, height, tick_label=labels, width=0.6,)
+            plt.xticks(rotation=60)
+            plt.gcf().subplots_adjust(bottom=0.3)
+            fname = f"plot_{metric.abbreviation}_{level}.png"
+            plt.savefig(f"data/analysis/levels/{fname}")
+            plt.cla()
+
+
 def _make_plot(metric: str, experiment: Experiment, short_name: str = "", sort_revert: bool = False):
     a1 = experiment.results.filter(
         metrics=[metric], levels=["A1"]).mean(metric)
@@ -94,14 +117,15 @@ def _make_plot(metric: str, experiment: Experiment, short_name: str = "", sort_r
 
 
 def plot_results(metrics: List[Metric], experiment: Experiment):
-    # for metric in metrics:
-    #     for topic in TOPICS:
-    #         _make_topic_metric_plot(
-    #             metric.name, topic, experiment, short_name=metric.abbreviation)
-    _make_plot("Words in Top 3000", experiment,
-               short_name="WIT", sort_revert=True)
+    _plot_levels(metrics, experiment)
+    for metric in metrics:
+        for topic in TOPICS:
+            _make_topic_metric_plot(
+                metric.name, topic, experiment, short_name=metric.abbreviation)
+    # _make_plot("Words in Top 3000", experiment,
+    #            short_name="WIT", sort_revert=True)
 
     # for metric in metrics:
     #     _make_plot(metric.name, experiment, short_name=metric.abbreviation)
 
-    _make_table(metrics, experiment)
+    # _make_table(metrics, experiment)

### 2023-03-29 src/classifier/load_data.py
play around with classifier

@@ -24,7 +24,7 @@ def load_glove_data():
     return word2idx, idx2word, np.array(embs)
 
 
-def get_train_test_ids(train_limit: int = 200_000):
+def get_train_test_ids(train_limit: int = 200_000, test_limit: int = 20_000):
     train, test = [], []
     with Session(engine) as session:
         q_simple_train = session.query(ClassifierDataset).filter_by(
@@ -32,9 +32,9 @@ def get_train_test_ids(train_limit: int = 200_000):
         q_standard_train = session.query(ClassifierDataset).filter_by(
             level="standard", partition="train").limit(train_limit)
         q_simple_test = session.query(ClassifierDataset).filter_by(
-            level="simple", partition="test")
+            level="simple", partition="test").limit(test_limit)
         q_standard_test = session.query(ClassifierDataset).filter_by(
-            level="standard", partition="test")
+            level="standard", partition="test").limit(test_limit)
         for item in q_simple_train:
             train.append((item.id, "simple"))
         for item in q_standard_train:

### 2023-03-31 write_dataset_csv.py
update dataset and classifier jupyter

@@ -1,0 +1,29 @@
+import csv
+from src.sql import engine
+from src.sql.models import ClassifierDataset
+from sqlalchemy.orm.session import Session
+
+
+def main():
+    train_rows = []
+    test_rows = []
+    train_csv_path = "data/dataset_train.csv"
+    test_csv_path = "data/dataset_test.csv"
+    with Session(engine) as session:
+        q = session.query(ClassifierDataset)
+        for item in q:
+            row = [item.id, item.level, item.text]
+            if item.partition == "train":
+                train_rows.append(row)
+            else:
+                test_rows.append(row)
+    with open(train_csv_path, "w", encoding="utf-8", newline="") as f:
+        writer = csv.writer(f)
+        writer.writerows(train_rows)
+    with open(test_csv_path, "w", encoding="utf-8", newline="") as f:
+        writer = csv.writer(f)
+        writer.writerows(test_rows)
+
+
+if __name__ == '__main__':
+    main()

### 2023-03-31 write_dataset_csv.py
update dataset writer

@@ -2,6 +2,7 @@ import csv
 from src.sql import engine
 from src.sql.models import ClassifierDataset
 from sqlalchemy.orm.session import Session
+import random
 
 
 def main():
@@ -11,12 +12,23 @@ def main():
     test_csv_path = "data/dataset_test.csv"
     with Session(engine) as session:
         q = session.query(ClassifierDataset)
+        train_rows_standard = []
+        train_rows_simple = []
+
         for item in q:
             row = [item.id, item.level, item.text]
             if item.partition == "train":
-                train_rows.append(row)
+                if item.level == "simple":
+                    train_rows_simple.append(row)
+                elif item.level == "standard":
+                    train_rows_standard.append(row)
             else:
                 test_rows.append(row)
+
+        random.shuffle(train_rows_standard)
+        train_rows.extend(train_rows_simple)
+        train_rows.extend(train_rows_standard[:len(train_rows_simple)])
+
     with open(train_csv_path, "w", encoding="utf-8", newline="") as f:
         writer = csv.writer(f)
         writer.writerows(train_rows)

### 2023-03-31 paper/paper.tex
finalize paper

@@
 7 commonly used complexity metrics have been employed, targeting chiefly lexis and syntax.
-In the limited scope of this experiment, GPT-3 is observed to at least partially aware of what constitutes reading difficulty.
+In the limited scope of this experiment, GPT-3 is observed to be at least partially aware of what constitutes reading difficulty.
 \end{abstract}
@@
 
-The feautre "Words in Top 3000 English words" uses unigram frequencies from the Google Web Trillion Word corpus\footnote{\url{https://ai.googleblog.com/2006/08/all-our-n-gram-are-belong-to-you.html}} to isolate the 3000 most frequent words of the English language.
+The feature "Words in Top 3000 English words" uses unigram frequencies from the Google Web Trillion Word corpus\footnote{\url{https://ai.googleblog.com/2006/08/all-our-n-gram-are-belong-to-you.html}} to isolate the 3000 most frequent words of the English language.
 Many research approaches to readability employ similar ideas to gain insight into the lexical complexity of texts, such as the BIV (basic Italian vocabulary) employed by \citep{dellorletta-etal-2011-read}.
@@
 
-Interesting discussions can be had around topics as well, since, depending on GPT-3's ability to generalize across domains, certain areas might be problem areas for it.
+Interesting discussions can be had around topics as well, since, depending on GPT-3's ability to generalize across domains, certain areas might constitute problem for it.
 For example, "semantics" and "quantum mechanics" would be hard to write about for A1 level students as they are likely to be children who have not interacted with those fields yet.
@@
 
-This work attempted to gauge true readability levels of GPT-3 output texts with prompts across 10 topics and CEFR proficiency levels, optimising thus for the second language learning application environment.
-Outputs computed from the metrics set suggest that GPT-3 is at least somewhat aware of syntax and lexis when performin generation for a target readability level.
+This work attempted to gauge true readability levels of GPT-3 output texts with prompts across 10 topics and CEFR proficiency levels, optimising for the second language learning application environment.
+Outputs computed from the metric set suggest that GPT-3 is at least somewhat aware of syntax and lexis when performin generation for a target readability level.
 

### 2023-04-02 paper/paper.tex
update paper

@@
 7 commonly used complexity metrics have been employed, targeting chiefly lexis and syntax.
+In addition, a classifier is also trained on Wikipedia data and used to evaluate the generated texts.
 In the limited scope of this experiment, GPT-3 is observed to be at least partially aware of what constitutes reading difficulty.
@@
 A number of complexity metrics are computed on the generated texts and compared with Simple Wikipedia\footnote{\url{https://simple.wikipedia.org/wiki/Main_Page}} as well as Wikipedia\footnote{\url{https://www.wikipedia.org/}} articles on the same topics in order to gauge the true readability levels of the model outputs.
-\section{Dataset and metrics}
+\section{Dataset and methods}
 \subsection{Dataset}
@@
 
+\subsection{Text classifier}
+
+A classifier was trained on Wikipedia and Simple Wikipedia data, specifically on a train dataset of approximately 240k paragraphs, evenly distributed across sourced from Wikipedia and Simple Wikipedia articles.
+A test set of 40k entries, sourced and distributed in the same manner as the train set, was also obtained to evaluate the data.
+The task of the classifier was to determine whether a given paragraph had originally been a part of Simple Wikipedia or (standard) Wikipedia - $P(standard)$ is the derived readability metric reported in the results.
+
+The model takes BERT sentence embeddings as input, which are fed through a single-layer LSTM with hidden size 126.
+LSTM outputs are then used to compute a distribution over the two labels in the dataset, namely "simple" (paragraph obtained from Simple Wikipedia) and "standard" (paragraph obtained from standard Wikipedia).
+Training was performed with the Adam optimizer and a batch size of 50 for 5 epochs.
+After evaluating the model at each epoch, it was observed that after the third epoch there are signs of overfitting, therefore the final parameters were those obtained after 3 epochs of training.
+Hardware-wise, the free tier of Google Colab was used - do to time limits set on the platform, several days were necessary for the full training process.
+After the third epoch, the model achieves 87.8\% macro-averaged $F_1$-score on the test set. While this is not state-of-the-art performance, it is likely good enough for the scope of this project.
+
 \section{Results}
+
+\subsection{Metrics}
 
@@
 
+\subsection{Classifier scores}
+
+\input{tables/classifier_scores.tex}
+
+Table~\ref{tab:classifier} shows classifier outputs, with highlighted values for each topic and difficulty level associated to lowest readability.
+The Wikipedia entries that were reported on in previous sections are not included here, as they may have been part of the train set of the classifier.
+
+These results show again that GPT-3 has learned what constitues readability, at least to some extent.
+For 4 of the topics presented in this study, the C2 prompt achieves the highest complexity (or lowest readability) level, with 3 more having B2 as the top entry.
+
+As was observed with the metrics, there tends to be good differentiation between the extremes (i.e.
+between A1 and C2 texts), but this is less true in the middle portions.
+Interestingly, prompts for C1 level texts often do not yield particularly difficult texts according to the classifier.
+Such a claim is confirmed by looking at the average values for each level, where C1 does score higher than A1 and A2, but lower than B2.
+Whether this is caused by this classifier's failure to capture some aspect of text readability, ot by GPT-3's lessened ability to generate material for this target level, remains to be confirmed by more robust experiments.
+
+On the matter of inherent topic complexities, there is unfortunately not much to talk about - no topic or set of topics emerge as consistently more difficult than others.
+However, \emph{Economics} and \emph{Semantics} do get the highest average score, indicating they are the least readable, while (more surprisingly) \emph{Color Blindness} is the most readable on average.
+
 \section{Discussion and future research}
@@
 Since recent research (e.g.
-\citealp{Deutsch_2020}) has also focused on readability assessment through neural approaches without such metrics, this could also be explored with AI text data.
-An attempt was made in relationship to this experiment to train a readability analyzer with GLoVe word embeddings \citep{pennington2014glove} and Wikipedia data, but this unfortunaly was inconclusive, and thus results from this analyzer haven't been reported.
+\citealp{Deutsch_2020}) has also focused on readability assessment through neural approaches without such metrics, this paper also attempted to adopt this strategy by training a classifier on Wikipedia and Simple Wikipedia data.
+Outputs derived from this model also support the theory of GPT-3 being at least somewhat aware of readability-related aspects of language.
+With regards to this section, it is nonetheless paramount to utilise a larger, SOTA readability model for exploring complexity - as the one employed here is only suitable for prototyping - as well as to perform the experiment on a richer repository of GPT-3 data.
 
-Mainly, however, future extensions of this experiment or related experiments would benefit from higher volume.
+In summary, future extensions of this experiment or related experiments would most benefit from higher volume.
 Simply put, more data is necessary across multiple domains to make educated conclusions about the quality of GPT-3 outputs in a teaching setting.
@@
 \section*{Acknowledgements}
+
+I thank Prof. Dr. Walt Detmar Meurers for exposing me to the topics of readability and automatic readability assessment, and for the excellent course he taught on the matter.
 
